% =============================================================================
% CHAPTER 3: METHODOLOGY
% =============================================================================

\chapter{METHODOLOGY}
\label{Chapter3}

% -----------------------------------------------------------------------------
\section{Introduction}
\label{sec:ch3intro}
% -----------------------------------------------------------------------------

This chapter outlines the method for the Dynamic Context-Aware Trie (DCA-Trie), a system aimed at solving the permissiveness problem in graph-constrained decoding. To check if DCA-Trie effectively balances structural faithfulness with contextual relevance, this chapter introduces a diagnostic metric called the Semantic Irrelevance Ratio (SIR). After defining the main problem and the SIR metric, the discussion explains the underlying semantic scoring method and the design of both DCA-Trie variants. Finally, it sets up the baseline systems, the evaluation protocol, and the specific ablation strategies used to isolate the framework's effects, establishing the limits for the experimental results discussed in Chapter~4.

% -----------------------------------------------------------------------------
\section{Formal Problem Specification}
\label{sec:ch3problem}
% -----------------------------------------------------------------------------

\subsection{Restating the Core Limitation of Existing Frameworks}

As established in Chapter~2, current graph-constrained frameworks (notably GCR \citep{luo2025-graph-constrained-reasoning} and DoG \citep{li2024-decoding-on-graphs}) rely on a deterministic oracle that looks only at the fixed graph $\mathcal{G}$ and the starting entities $\mathcal{E}_q$. Regardless of whether the system builds this oracle before decoding or expands it on the fly, the set of valid tokens at any step $t$ remains locked to structural logic:
\begin{equation}
  \mathcal{W}_{\text{val}}^{\text{GCR}}(t) = f(\mathcal{G},\, \mathcal{E}_q)
  \quad \forall\, t
  \label{eq:gcr_oracle}
\end{equation}

By ignoring both the semantic intent of the question $q$ and the reasoning trajectory already captured in the prefix $y_{<t}$, this formulation treats every structurally reachable path as equally plausible. In dense, multi-hop Freebase environments, this blindness causes catastrophic search space explosion. The framework might admit over 8,000 paths at depth 3, even though only a single path logically answers the query.

The core issue is that these systems conflate two separate requirements. \textit{Structural validity} ensures the LLM only generates tokens that correspond to real graph edges, which current models do well. \textit{Contextual relevance} ensures the LLM only generates tokens that logically connect to the specific question at hand, which current models fail to do. This gap between structural compliance and semantic usefulness is the \textit{permissiveness problem}.

\subsection{The Upgraded Oracle Specification}

DCA‑Trie addresses this problem by defining an upgraded constraint oracle that combines structural validity with semantic context. The oracle conditions on both the input question and the evolving generation prefix:
\begin{equation}
  \mathcal{W}_{\text{val}}^{\text{DCA}}(t) = f(\mathcal{G},\, \mathcal{E}_q,\,
  q,\, y_{<t})
  \label{eq:dca_oracle}
\end{equation}
where $q$ is the natural language query and $y_{<t} = (y_1, \ldots, y_{t-1})$ is the generated prefix at step $t$. To be methodologically sound, the upgraded oracle must satisfy three conditions aligned with the project objectives in Chapter~1:

\begin{enumerate}
  \item \textbf{Structural Faithfulness (Condition~F):} Every candidate token $v \in \mathcal{W}_{\text{val}}^{\text{DCA}}(t)$ must correspond to a valid prefix of a path $p \in \mathcal{G}$ at every step $t$. This preserves the 100\% structural faithfulness guarantee achieved by GCR and DoG and prevents ungrounded tokens from entering the valid set.

  \item \textbf{Semantic Relevance Reduction (Condition~R):} The Semantic Irrelevance Ratio (SIR) under DCA‑Trie must be lower than the corresponding SIR under GCR when averaged over the evaluation corpus. Using SIR as defined in Section~\ref{sec:ch3sir}:
        \begin{equation}
          \overline{\text{SIR}}\!\left(\mathcal{W}^{\text{DCA}}\right) <
          \overline{\text{SIR}}\!\left(\mathcal{W}^{\text{GCR}}\right)
          \label{eq:sir_condition}
        \end{equation}

  \item \textbf{Recall Preservation (Condition~P):} Semantic pruning must not remove gold paths excessively. Operationally, the false negative rate (FNR) on the WebQSP validation split must remain below 5\%:
        \begin{equation}
          \text{FNR} = \frac{|\{q \in \mathcal{Q}_{\text{val}} :
          p^*_q \notin \mathcal{W}_{\text{val}}^{\text{DCA}}(t)\}|}{|\mathcal{Q}_{\text{val}}|}
          < 0.05
          \label{eq:fnr}
        \end{equation}
        where $p^*_q$ is the gold path for question $q$ and $\mathcal{Q}_{\text{val}}$ is the held-out 100-question validation set.
\end{enumerate}


% -----------------------------------------------------------------------------
\section{System Architecture Overview}
\label{sec:ch3arch}
% -----------------------------------------------------------------------------

DCA-Trie focuses on the constraint oracle layer, so its architecture directly builds on the baseline pipeline from Graph-Constrained Reasoning (GCR) \citep{luo2025-graph-constrained-reasoning}. This reuse helps isolate the experimental effects. By keeping the rest of the generation pipeline constant, the system ensures that any improvement in latency or search efficiency comes only from the new oracle mechanism. The four layers work as follows:

\begin{enumerate}
  \item \textbf{Entity Linking Layer (Unmodified):} A named entity recognition tool first identifies the core query entities $\mathcal{E}_q$ from the raw text question $q$. These entities dictate where the downstream search algorithm can start exploring the knowledge graph.

  \item \textbf{Constraint Oracle Layer (DCA-Trie Contribution):} This layer controls which tokens the LLM can generate. In GCR, this layer creates a static prefix tree (a KG-Trie) that includes every possible blind path reachable within $L$ hops of the starting entities. DCA-Trie completely revamps this. Through either static pre-filtering (v1) or dynamic, step-by-step expansion (v2), the new oracle ensures that no structural edge is added to the trie unless it also shows semantic relevance to the question.

  \item \textbf{Constrained Decoding Layer (Unmodified):} During autoregressive generation using beam search, this layer intercepts the LLM logit distribution and applies a binary mask retrieved from the oracle layer. By forcing the LLM to output only tokens that are structurally and semantically approved, it maintains 100\% ground-truth faithfulness to the knowledge graph.

  \item \textbf{Inductive Reasoning Layer (Unmodified):} After the constrained beam search produces $K$ top reasoning paths, GPT-4o-mini analyzes these paths at the same time to synthesize the final natural language answer.
\end{enumerate}

Figure~\ref{fig:architecture} shows the layout, isolating the constraint oracle layer to emphasize where the core semantic intervention takes place.

% \begin{figure}[H]
%   \centering
%   \begin{tikzpicture}[
%     node distance=0.6cm and 1.2cm,
%     box/.style={draw, rectangle, rounded corners=4pt, minimum height=1.0cm,
%         minimum width=3.5cm, align=center, thick, font=\small},
%     contrib/.style={draw, rectangle, rounded corners=4pt, minimum height=1.0cm,
%         minimum width=3.5cm, align=center, thick, font=\small,
%         fill=teal!12, draw=teal!60!black},
%     arrow/.style={-{Stealth[scale=1.1]}, thick},
%     label/.style={font=\footnotesize\itshape, text=gray}
%     ]
%     \node[box]    (el)   {Entity Linking};
%     \node[contrib, below=of el] (oracle) {Constraint Oracle\\(DCA‑Trie)};
%     \node[box,    below=of oracle] (cd) {Constrained Decoding\\(Beam Search, $K$ beams)};
%     \node[box,    below=of cd]    (ir) {Inductive Reasoning\\(GPT‑4o‑mini)};
%     \node[box,    below=of ir]    (ans){Final Answer};

%     \draw[arrow] (el)     -- node[label, right] {$\mathcal{E}_q$} (oracle);
%     \draw[arrow] (oracle) -- node[label, right] {Mask $\mathbf{m}_t$} (cd);
%     \draw[arrow] (cd)     -- node[label, right] {$K$ reasoning paths} (ir);
%     \draw[arrow] (ir)     -- (ans);

%     \node[label, right=0.3cm of oracle] {\textit{← Primary contribution}};
%   \end{tikzpicture}
%   \caption[DCA‑Trie system architecture]{DCA‑Trie system architecture. The constraint oracle layer (teal) is the primary contribution; all other layers are inherited from GCR~\citep{luo2025-graph-constrained-reasoning}.}
%   \label{fig:architecture}
% \end{figure}

\begin{figure}[H]
  \centering
  \resizebox{\textwidth}{!}{%
    \begin{tikzpicture}[
        >=stealth, font=\sffamily\small,
        kgnode/.style={draw=gray!70,fill=blue!8,rounded corners=3pt,inner sep=4pt,font=\sffamily\scriptsize,align=center},
        pathbox/.style={draw=gray!50,fill=gray!5,rounded corners=4pt,inner sep=6pt,minimum width=5.0cm,align=left,font=\sffamily\scriptsize},
        filtpath/.style={draw=gray!25,fill=gray!3,rounded corners=4pt,inner sep=6pt,minimum width=5.0cm,align=left,font=\sffamily\scriptsize,text=gray!45},
        ansbox/.style={draw=teal!60!black,fill=teal!8,rounded corners=4pt,inner sep=8pt,minimum width=3.0cm,align=left,font=\sffamily\scriptsize},
        stepbox/.style={draw=black!40,fill=white,rounded corners=3pt,inner sep=5pt,minimum width=2.6cm,minimum height=0.85cm,align=center,font=\sffamily\scriptsize},
        dcabox/.style={draw=teal!70!black,fill=teal!10,rounded corners=3pt,inner sep=5pt,minimum width=2.6cm,minimum height=0.85cm,align=center,font=\sffamily\scriptsize\bfseries},
        llmcirc/.style={circle,draw=gray!60,fill=blue!5,minimum size=1.1cm,align=center,font=\sffamily\scriptsize},
        bigllm/.style={circle,draw=teal!60,fill=teal!8,minimum size=1.25cm,align=center,font=\sffamily\scriptsize},
        arr/.style={->,thick,draw=gray!70},
        rarr/.style={->,thick,draw=red!60},
        tarr/.style={->,thick,draw=teal!70!black},
      ]

      % Knowledge Graph Section
      \node[font=\sffamily\footnotesize\bfseries]at(2.5,0.2){Knowledge Graph};
      \node[kgnode](usa)at(0,-2.5){USA};
      \node[kgnode](gwb)at(2.1,-1.0){George W.\\Bush};
      \node[kgnode](lb)at(4.5,-1.0){Laura\\Bush};
      \node[kgnode](bo)at(2.3,-2.5){Barack\\Obama};
      \node[kgnode](mo)at(4.5,-2.5){Michelle\\Obama};
      \node[kgnode](dt)at(2.3,-4.0){Donald\\Trump};
      \node[kgnode](mt)at(4.5,-4.0){Melania\\Trump};

      \draw[arr](usa)--node[above,sloped,font=\sffamily\tiny,pos=0.45]{Ex-president}(gwb);
      \draw[arr](usa)--node[above,font=\sffamily\tiny,pos=0.5]{Ex-president}(bo);
      \draw[arr](usa)--node[below,sloped,font=\sffamily\tiny,pos=0.45]{President}(dt);
      \draw[arr](gwb)--node[above,font=\sffamily\tiny]{Spouse\_of}(lb);
      \draw[arr](bo)--node[above,font=\sffamily\tiny]{Spouse\_of}(mo);
      \draw[arr](dt)--node[above,font=\sffamily\tiny]{Marry\_to}(mt);

      \node[draw=red!40,fill=red!5,rounded corners=4pt,inner sep=6pt,align=left,
        font=\sffamily\scriptsize,minimum width=2.9cm]
      (qbox)at(2.5,-6.4){\textbf{Q:} Who is the spouse of\\the ex-president of USA?};
      \node[circle,draw=black!35,fill=black!6,font=\sffamily\footnotesize\bfseries,inner sep=3pt]at(-0.1,-5.1){1};
      \node[font=\sffamily\scriptsize,text=black!65,align=center]at(1.5,-5.1){KG-Trie Construction};

      % DCA-Trie Constraint Section
      \draw[draw=teal!60!black,dashed,rounded corners=6pt,line width=1.2pt](6.0,0.3)rectangle(10.5,-7.4);
      \node[font=\sffamily\footnotesize\bfseries,text=teal!70!black]at(8.25,0.1){DCA-Trie Constraint};
      \node[dcabox,minimum width=3.8cm](vf)at(8.25,-0.8){Semantic Filtering\\score$(p,q,y_{<t})\geq\tau$};
      \node[draw=teal!45,fill=teal!5,rounded corners=4pt,minimum width=4.0cm,minimum height=2.4cm](triebox)at(8.25,-2.9){};
      \node[font=\sffamily\scriptsize\bfseries,text=teal!80!black]at(8.25,-1.9){Filtered KG-Trie};

      \node[circle,draw=teal!60,fill=teal!18,inner sep=2pt,font=\sffamily\tiny](tr)at(8.25,-2.3){USA};
      \node[circle,draw=teal!50,fill=white,inner sep=1.5pt,font=\sffamily\tiny](tn1)at(7.25,-2.9){GWB};
      \node[circle,draw=teal!50,fill=white,inner sep=1.5pt,font=\sffamily\tiny](tn2)at(8.25,-2.9){BO};
      \node[circle,draw=gray!25,fill=gray!4,inner sep=1.5pt,font=\sffamily\tiny,text=gray!40](tn3)at(9.25,-2.9){DT};
      \node[circle,draw=teal!50,fill=white,inner sep=1.5pt,font=\sffamily\tiny](tl1)at(7.0,-3.5){LB};
      \node[circle,draw=teal!50,fill=white,inner sep=1.5pt,font=\sffamily\tiny](tl2)at(8.5,-3.5){MO};

      \draw[teal!50,thin](tr)--(tn1); \draw[teal!50,thin](tr)--(tn2);
      \draw[gray!25,thin,dashed](tr)--(tn3);
      \draw[teal!50,thin](tn1)--(tl1); \draw[teal!50,thin](tn2)--(tl2);
      \node[font=\sffamily\tiny,text=gray!50,align=center]at(9.25,-3.5){pruned\\$<\!\tau$};

      \node[draw=teal!35,fill=teal!5,rounded corners=3pt,inner sep=4pt,font=\sffamily\tiny,align=center]
      (upd)at(8.25,-4.9){v2: step-wise update\\re-score $\mathcal{N}(e^{*})$ vs $(q,y_{<t})$};
      \node[llmcirc](kgllm)at(8.25,-6.5){KG-spec.\\LLM};

      \draw[rarr,rounded corners](qbox.east)-- (5.6,-6.4) |- (vf.west);
      \draw[tarr](vf)--(triebox.north);
      \draw[tarr](triebox.south)--(upd.north);
      \draw[tarr](upd)--(kgllm);
      \node[circle,draw=teal!70!black,fill=teal!10,font=\sffamily\footnotesize\bfseries,inner sep=3pt]at(6.5,-7){2};

      % Reasoning Paths and Hypothesis Answers Section
      \node[font=\sffamily\footnotesize\bfseries,align=center]at(13.5,0.2){Reasoning Paths and\\Hypothesis Answers};
      \node[pathbox](p1)at(13.7,-1.3){%
        \texttt{\# Reasoning Path (t=1):}\\[1pt]
        USA $\to$ G.W.\,Bush $\to$ Laura Bush\\[3pt]
        \texttt{\# Answer: Laura Bush}};
      \node[pathbox](p2)at(13.7,-3.2){%
        \texttt{\# Reasoning Path (t=2):}\\[1pt]
        USA $\to$ B.\,Obama $\to$ Michelle Obama\\[3pt]
        \texttt{\# Answer: Michelle Obama}};
      \node[filtpath](p3)at(13.7,-5.5){%
        \texttt{[Excluded by DCA-Trie]}\\[1pt]
        USA $\to$ D.\,Trump $\to$ Melania Trump\\[3pt]
        score $<\tau$: structurally valid,\\
        semantically irrelevant};

      \draw[tarr,rounded corners](kgllm.east)-- (10.8,-6.5) |- (p1.west);
      \draw[tarr,rounded corners](kgllm.east)-- (10.8,-6.5) |- (p2.west);
      \draw[tarr,dashed,draw=teal!60,rounded corners](kgllm.east)-- (10.8,-6.5) |- (p3.west);

      % General LLM Section
      \node[font=\sffamily\footnotesize\bfseries]at(18.5,0.2){General LLM};
      \node[bigllm](gllm)at(18.7,-1.8){General\\LLM};
      \node[ansbox](ans)at(18.7,-4.8){%
        \textbf{A:} Based on the paths,\\
        the answers are:\\[3pt]
        \textbf{Laura Bush,}\\
        \textbf{Michelle Obama.}\\[4pt]
        \small\textit{(Trump path excluded}\\
        \textit{by semantic filter)}};
      \node[circle,draw=black!35,fill=black!6,font=\sffamily\footnotesize\bfseries,inner sep=3pt]at(16.9,-0.8){3};
      \node[font=\sffamily\scriptsize,text=black!65]at(18.5,-0.8){Inductive Reasoning};

      \draw[arr,rounded corners](p1.east) -| (16.4,-1.8) -- (gllm.west);
      \draw[arr,rounded corners](p2.east) -| (16.4,-1.8) -- (gllm.west);
      \draw[tarr](gllm)--(ans);

      % Footer Steps Section
      \draw[gray!50,dashed](0,-7.8)--(20.0,-7.8);
      \node[stepbox]at(2.5,-8.6){1.\ KG-Trie Construction\\BFS from $\mathcal{E}_q$};
      \node[dcabox]at(8.25,-8.6){2.\ DCA-Trie Filtering\\score$(p,q,y_{<t})\geq\tau$};
      \node[stepbox]at(13.5,-8.6){Constrained Beam Search\\$K$ hypothesis paths};
      \node[stepbox]at(17.5,-8.6){3.\ Inductive Reasoning\\synthesise answer};

      \draw[arr](4,-8.6)--(6.8,-8.6);
      \draw[tarr](9.7,-8.6)--(11.8,-8.6);
      \draw[arr](15.2,-8.6)--(16.0,-8.6);
    \end{tikzpicture}
  }%
  \caption[DCA‑Trie System Architecture]{DCA‑Trie System Architecture. The constraint oracle layer (teal) is the primary contribution; all other layers are inherited from GCR~\citep{luo2025-graph-constrained-reasoning}.}
  \label{fig:architecture}
\end{figure}

% -----------------------------------------------------------------------------
\section{The Semantic Irrelevance Ratio: Definition and Measurement}
\label{sec:ch3sir}
% -----------------------------------------------------------------------------

\subsection{Standard Metric Deficiencies}

Current performance markers for knowledge graph decoding, such as Hits@1, F1, and structural faithfulness ratios, have a significant drawback: they focus on output instead of the process. While these metrics show whether a framework found the correct answer and stayed within the graph, they do not reveal how the framework reached that answer. A model may solve the correct gold path, but it could wander through a vast number of irrelevant paths that waste resources.

Therefore, confirming accuracy is not enough; we need to measure the constraint mechanism itself to ensure computational focus. To assess how lenient the oracle is, without considering the final output accuracy, this thesis presents a new diagnostic metric: the Semantic Irrelevance Ratio (SIR).

\subsection{Formal Definition}

Suppose $\mathcal{P}(q, t)$ represents all paths that the constraint oracle allows at decoding step $t$ for question $q$. To calculate SIR, the system must first define irrelevance. An individual candidate path $p \in \mathcal{P}(q, t)$ is structurally valid, but it fails the semantic stringency test if its contextual similarity drops below a hard baseline threshold $\tau_{\text{ref}}$:
\begin{equation}
  \text{irrelevant}(p, q, y_{<t}) = \mathbf{1}\!\left[
    \cos\!\bigl(E(p),\; E(\text{concat}(q, y_{<t}))\bigr) < \tau_{\text{ref}}
    \right]
  \label{eq:irrelevant}
\end{equation}
where $E(\cdot)$ is the \texttt{all-MiniLM-L6-v2} sentence encoder, and $\tau_{\text{ref}} = 0.3$ is a fixed reference threshold. This reference threshold is intentionally non-tunable and separate from DCA‑Trie's admission threshold $\tau$.

The step-level SIR for question $q$ at decoding step $t$ is then defined as:
\begin{equation}
  \text{SIR}(q, t) = \frac{\sum_{p \in \mathcal{P}(q,t)}
    \text{irrelevant}(p, q, y_{<t})}{|\mathcal{P}(q, t)|}
  \label{eq:sir_step}
\end{equation}

Corpus-level SIR is computed by averaging over all questions and all valid decoding steps up to depth $L$:
\begin{equation}
  \overline{\text{SIR}} = \frac{1}{|\mathcal{Q}|} \sum_{q \in \mathcal{Q}}
  \frac{1}{L_q} \sum_{t=1}^{L_q} \text{SIR}(q, t)
  \label{eq:sir_aggregate}
\end{equation}
where $L_q$ is the hop length of the generated reasoning chain for question $q$.

\subsection{Properties and Interpretation}

By definition, $\overline{\text{SIR}} \in [0,1]$. A value near 1 indicates that most admitted paths are irrelevant, so the model still faces a large and noisy search space. A value near 0 indicates that the constraint set is tightly aligned with question semantics. Prior constrained frameworks such as GCR and DoG tend to show high SIR, especially at larger hop depths where path counts increase rapidly. DCA‑Trie is designed to reduce SIR by filtering semantically weak paths while retaining structurally valid ones. In reporting, SIR is analysed by hop depth to evaluate how this effect scales with reasoning complexity.

% -----------------------------------------------------------------------------
\section{Semantic Relevance Scoring Mechanism}
\label{sec:ch3scorer}
% -----------------------------------------------------------------------------

\subsection{Scorer Architecture}

The semantic relevance scorer implements the transition from a static oracle $f(\mathcal{G}, \mathcal{E}_q)$ to a context-aware oracle $f(\mathcal{G}, \mathcal{E}_q, q, y_{<t})$. It takes a candidate KG path $p$, the question $q$, and the partial generation $y_{<t}$, and returns a relevance score in $[0,1]$.

A path $p = (e_0, r_1, e_1, \ldots, r_L, e_L)$ is first serialized into a text sequence:
\begin{equation}
  \text{str}(p) = e_0 \;\|\; r_1 \;\|\; e_1 \;\|\; \cdots \;\|\; r_L \;\|\; e_L
  \label{eq:pathstr}
\end{equation}
where entity IDs are mapped to readable Freebase names when available, and relation strings are normalized (e.g., underscore replacement). The query-side context is serialized as:
\begin{equation}
  \text{query}(q, y_{<t}) = q \;\|\; \texttt{[SEP]} \;\|\;
  \text{str}(y_{<t})
  \label{eq:querystr}
\end{equation}
where $\text{str}(y_{<t})$ is the detokenised generation prefix.

Both $\text{str}(p)$ and $\text{query}(q, y_{<t})$ are encoded using \texttt{all-MiniLM-L6-v2} \citep{reimers2019sbert} into 384-dimensional vectors. Relevance is computed as cosine similarity:
\begin{equation}
  \text{score}(p,\, q,\, y_{<t}) = \cos\!\bigl(
  E\!\left(\text{str}(p)\right),\;\;
  E\!\left(\text{query}(q, y_{<t})\right)
  \bigr)
  \label{eq:scorer}
\end{equation}

A candidate path is admitted into $\mathcal{W}_{\text{val}}^{\text{DCA}}(t)$ if and only if $\text{score}(p, q, y_{<t}) \geq \tau$, where $\tau$ is the admission threshold. This binary decision integrates the semantic score into the same logit-masking framework defined in Equation. The full scoring workflow is shown in Figure~\ref{fig:scorer}.

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[
    box/.style={draw, rounded corners, minimum width=2.5cm, minimum height=0.8cm, align=center, font=\small},
    encoder/.style={draw, fill=blue!10, rounded corners, minimum width=2.5cm, minimum height=0.8cm, align=center, font=\small},
    arrow/.style={-{Stealth[scale=1]}, thick},
    node distance=1.5cm
    ]

    % Inputs
    \node[box] (query) {Question $q$ \\ + Context $y_{<t}$};
    \node[box, right=2cm of query] (path) {Candidate Path $p$};

    % Formatter
    \node[box, below=1cm of query] (qformat) {String Formatter\\(Eq 3.9)};
    \node[box, below=1cm of path] (pformat) {String Formatter\\(Eq 3.8)};

    % Encoders
    \node[encoder, below=1cm of qformat] (qenc) {Sentence Encoder\\(MiniLM)};
    \node[encoder, below=1cm of pformat] (penc) {Sentence Encoder\\(MiniLM)};

    % Cosine
    \node[box, diamond, aspect=2, below=1cm of $(qenc)!0.5!(penc)$] (cos) {Cosine\\Similarity};

    % Threshold
    \node[box, below=1cm of cos] (thresh) {Admit if \\ $s \geq \tau$};

    % Connections
    \draw[arrow] (query) -- (qformat);
    \draw[arrow] (path) -- (pformat);

    \draw[arrow] (qformat) -- node[left, font=\footnotesize] {Query str} (qenc);
    \draw[arrow] (pformat) -- node[right, font=\footnotesize] {Path str} (penc);

    \draw[arrow] (qenc) -- node[left, font=\footnotesize] {Vector $\mathbf{u}$} (cos);
    \draw[arrow] (penc) -- node[right, font=\footnotesize] {Vector $\mathbf{v}$} (cos);

    \draw[arrow] (cos) -- node[right, font=\footnotesize] {Score $s$} (thresh);
  \end{tikzpicture}
  \caption[Semantic Relevance Scoring Mechanism]{Semantic Relevance Scoring Mechanism.}
  \label{fig:scorer}
\end{figure}

\subsection{Threshold Selection}

The admission threshold $\tau$ is tuned on a separate 100-question WebQSP validation split. Selection prioritizes Condition~P (low false negatives) while still maximizing trie reduction relative to GCR. We sweep $\tau \in [0.1, 0.6]$ with step size 0.05 and select the highest value that satisfies $\text{FNR} < 0.05$. Thresholds are tuned independently for v1 and v2, since v2 uses step-wise context updates and therefore follows a different pruning dynamic.

\subsection{Encoder Caching}

To reduce repeated computation during beam search, path embeddings $E(\text{str}(p))$ are cached. In v1, caching is done up front because the trie is fixed after construction. In v2, embeddings are cached lazily as new branches are expanded. In both cases, caching avoids redundant encoder calls for previously seen paths. By contrast, the query-context embedding $E(\text{query}(q, y_{<t}))$ must be recomputed whenever the generation prefix changes.

% -----------------------------------------------------------------------------
\section{DCA‑Trie v1: Static Semantic Filtering}
\label{sec:ch3v1}
% -----------------------------------------------------------------------------

\subsection{Design Rationale}

DCA‑Trie v1 addresses the permissiveness problem while keeping the efficient static-trie design used in GCR. The key idea is to apply semantic filtering once, before decoding begins, using only the input question $q$. This is useful because many paths are structurally reachable but semantically unrelated to the question. Scoring paths against $q$ during preprocessing removes weak candidates early and reduces the search space before autoregressive generation starts.

Methodologically, this changes the oracle from $f(\mathcal{G}, \mathcal{E}_q)$ to $f(\mathcal{G}, \mathcal{E}_q, q)$ without changing generation-time lookup mechanics. The decoder still queries a precompiled trie, so local lookup remains $O(d)$ as in GCR. Figure~\ref{fig:v1} summarises this offline filtering workflow.

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[
    box/.style={draw, rounded corners, minimum width=2.9cm, minimum height=0.95cm, align=center, font=\small},
    db/.style={draw, cylinder, shape border rotate=90, aspect=0.25, minimum width=2.9cm, minimum height=1.25cm, align=center, font=\small, fill=gray!10},
    arrow/.style={-{Stealth[scale=1]}, thick},
    node distance=1.0cm
    ]

    \node[box] (query) {Question $q$};
    \node[db, right=2.5cm of query] (kg) {Knowledge\\Graph $\mathcal{G}$};

    \node[box, below=0.95cm of query] (qenc) {Precompute\\Query Vector $\mathbf{u}$};
    \node[box, below=0.95cm of kg] (bfs) {BFS Enumeration\\(Max $L$ hops)};

    \node[box, below=0.95cm of bfs] (paths) {Candidate Paths $\mathcal{P}$};

    \node[box, fill=teal!10, below=1.45cm of paths, xshift=-2.2cm, text width=8.2cm, align=left, minimum height=3.0cm] (filter) {\centering\textbf{Semantic Filter Loop}\\[2pt]
      \raggedright For each $p \in \mathcal{P}$:\\
      1. Compute $\mathbf{v}_p \leftarrow E(\text{str}(p))$\\
      2. Compute $s_p \leftarrow \cos(\mathbf{u}, \mathbf{v}_p)$\\
      3. Keep $p$ if $s_p \geq \tau$};

    \node[box, below=1.1cm of filter] (trie) {Construct Filtered\\KG‑Trie $\mathcal{T}^{\text{v1}}$};

    \draw[arrow] (query) -- (qenc);
    \draw[arrow] (kg) -- (bfs);
    \draw[arrow] (bfs) -- (paths);

    \draw[arrow] (qenc.south) -- ([xshift=-3.2cm]filter.north);
    \draw[arrow] (paths.south) -- ([xshift=2.2cm]filter.north);

    \draw[arrow] (filter) -- node[right, font=\footnotesize] {$\mathcal{P}_{\text{filtered}}$} (trie);
  \end{tikzpicture}
  \caption[Static Semantic Filtering Pipeline (DCA‑Trie v1)]{Static Semantic Filtering Pipeline for DCA‑Trie v1.}
  \label{fig:v1}
\end{figure}

\subsection{Algorithm}

\begin{figure}[H]
  \centering
  \begin{algorithm}[H]
    \caption{DCA‑Trie v1: Static Semantic Filtering at Construction Time}
    \label{alg:v1}
    \begin{algorithmic}[1]
      \Require Knowledge graph $\mathcal{G}$; question entities $\mathcal{E}_q$;
      question $q$; max hop depth $L$; admission threshold $\tau$;
      sentence encoder $E$
      \Ensure Semantically filtered KG‑Trie $\mathcal{T}^{\text{v1}}$

      \State \textbf{Precompute query embedding:}
      $\mathbf{u} \gets E(\text{query}(q, \varepsilon))$
      \Comment{$y_{<t} = \varepsilon$ at construction time}

      \State \textbf{Enumerate paths:}
      $\mathcal{P} \gets \textsc{BFS}(\mathcal{G}, \mathcal{E}_q, L)$
      \Comment{All paths within $L$ hops, as in GCR}

      \State \textbf{Score and filter:}
      \For{each path $p \in \mathcal{P}$}
      \State $\mathbf{v}_p \gets E(\text{str}(p))$
      \Comment{Precompute and cache path embedding}
      \State $s_p \gets \cos(\mathbf{u}, \mathbf{v}_p)$
      \If{$s_p \geq \tau$}
      \State Add $p$ to $\mathcal{P}_{\text{filtered}}$
      \EndIf
      \EndFor

      \State \textbf{Construct filtered trie:}
      $\mathcal{T}^{\text{v1}} \gets \textsc{BuildTrie}(\mathcal{P}_{\text{filtered}})$

      \State \Return $\mathcal{T}^{\text{v1}}$

      \Statex
      \Statex \textbf{At generation time} (identical to GCR):
      \For{each step $t$}
      \State $\mathcal{V}_t \gets \textsc{TrieLookup}(\mathcal{T}^{\text{v1}}, y_{<t})$
      \State $\mathbf{m}_t[v] \gets \mathbf{1}[v \in \mathcal{V}_t]$ for all $v$
      \State $\tilde{\boldsymbol{\alpha}}_t \gets \mathbf{m}_t \odot
        \textsc{LLMLogits}(y_{<t}, x)$
      \State $y_t \gets \text{beam-search-step}(\tilde{\boldsymbol{\alpha}}_t)$
      \EndFor
    \end{algorithmic}
  \end{algorithm}
  \caption[Algorithm for DCA‑Trie v1]{DCA‑Trie v1 algorithm.}
  \label{fig:alg:v1}
\end{figure}

\subsection{Algorithm Description}

Algorithm~\ref{alg:v1} runs in two separate phases: an offline filtering phase and an online generation phase. To explain how the mechanism works, the process unfolds in the following sequence:

\begin{enumerate}
  \item \textbf{Precompute the Query State:} The system first embeds the raw question into a dense vector space, with an empty generation prefix at this stage.
  \item \textbf{Enumerate the Structural Space:} The system uses standard breadth-first search (BFS) to map out every possible graph path starting from the question entities, going up to a maximum depth $L$. This is similar to the baseline GCR approach.
  \item \textbf{Score and Prune:} This is where semantic intervention takes place. The system measures the cosine similarity between the question embedding and every path it has enumerated. Any path that scores below the threshold $\tau$ is discarded as contextually irrelevant.
  \item \textbf{Construct the Static Trie:} The remaining paths are compiled into a static prefix tree ($\mathcal{T}^{\text{v1}}$). This tree outlines the complete, semantically pruned valid search space.
  \item \textbf{Autoregressive Decoding (Online):} During generation, the LLM queries the pre-built trie at each token step. The trie returns a binary mask that limits the model to output only tokens that follow the pre-approved paths.
\end{enumerate}

This shows that semantic selectivity happens only once. By pruning the search space completely before generation starts, the decoding process stays fast, stable, and easily comparable to existing baselines.

\subsection{Complexity Analysis}

This section presents a complexity analysis of DCA‑Trie v1 at the implementation level, based on the data structure assumptions used in this project. Building $\mathcal{T}^{\text{v1}}$ has an offline cost of $O(|\mathcal{P}| \cdot d_E)$, where $|\mathcal{P}|$ is the number of BFS-enumerated paths and $d_E = 384$ is the embedding dimension. This is a one-time preprocessing cost. During decoding, the lookup complexity is $O(d)$, which matches GCR, because accessing the trie uses the same constant-time child lookup structure. Memory usage scales as $O(|\mathcal{P}_{\text{filtered}}| \cdot L)$, which is usually much smaller than GCR's $O(|\mathcal{P}| \cdot L)$ because semantic filtering removes irrelevant paths.

\subsection{Faithfulness Guarantee}

DCA‑Trie v1 maintains structural faithfulness by design. Every path kept in $\mathcal{P}_{\text{filtered}}$ comes from $\textsc{BFS}(\mathcal{G}, \mathcal{E}_q, L)$, ensuring that each triplet $(e_{i-1}, r_i, e_i)$ remains a valid graph edge. Semantic filtering only removes paths and does not add new tokens or edges. Therefore, any token passed through $\mathcal{T}^{\text{v1}}$ stays structurally grounded, meeting Condition~F when the token-mask construction and tokenizer alignment are correct.

% -----------------------------------------------------------------------------
\section{DCA‑Trie v2: Step‑Wise Dynamic Expansion}
\label{sec:ch3v2}
% -----------------------------------------------------------------------------

\subsection{Design Rationale}

DCA‑Trie v2 implements the full dynamic oracle in Equation~\eqref{eq:dca_oracle} by conditioning constraints on the evolving prefix $y_{<t}$. Unlike v1, it does not assume that question-level intent alone is sufficient throughout decoding. As generation progresses, $y_{<t}$ reveals the path decisions already made, so the relevant neighbourhood can change step by step. For example, after selecting \textit{Elvis\_Presley $\to$ starred\_in $\to$ Blue\_Hawaii}, plausible next expansions should focus on \textit{Blue\_Hawaii} rather than the initial entity set.

Architecturally, v2 follows the step-wise traversal pattern in DoG \citep{li2024-decoding-on-graphs} but adds semantic gating to each local expansion. After each committed entity $e_t$, the trie expands only to structurally valid neighbours that also satisfy the threshold $\tau$:
\begin{equation}
  \mathcal{T}_{t+1} = \mathcal{T}_t \;\cup\; \bigl\{(e_t,\, r,\, e') :
  (e_t, r, e') \in \mathcal{T}_{\mathcal{G}} \;\wedge\;
  \text{score}(e', q, y_{<t}) \geq \tau\bigr\}
  \label{eq:v2expansion}
\end{equation}
where $\mathcal{T}_{\mathcal{G}}$ is the set of graph triplets. The core difference from permissive dynamic expansion is therefore not when expansion happens, but how candidates are admitted: every step remains graph-valid and semantically filtered. The full workflow is shown in Figure~\ref{fig:v2}.

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[
    box/.style={draw, rounded corners, minimum width=2.5cm, minimum height=0.8cm, align=center, font=\small},
    db/.style={draw, cylinder, shape border rotate=90, aspect=0.25, minimum width=2.5cm, minimum height=1.2cm, align=center, font=\small, fill=gray!10},
    arrow/.style={-{Stealth[scale=1]}, thick},
    node distance=1.2cm
    ]

    \node[box] (step) {Decoding Step $t$\\with partial $y_{<t}$};
    \node[box, below=1cm of step] (llm) {LLM calculates\\logits for $y_t$};
    \node[box, right=1.5cm of step] (trie) {Current KG‑Trie $\mathcal{T}_{t-1}$};

    \node[box, below=1cm of trie] (lookup) {Lookup Valid\\Tokens $\mathcal{V}_t$};

    \node[box, fill=purple!10, below=1.5cm of $(llm)!0.5!(lookup)$] (mask) {Logit Masking\\(Eq 2.2)};

    \node[box, below=1cm of mask] (commit) {Commit Entity $e_t$};

    \node[box, fill=teal!10, text width=7cm, right=2.5cm of mask] (expand) {\textbf{Dynamic Semantic Expansion}\\1. $\mathbf{u}_t \leftarrow E(\text{query}(q, y_{\leq t}))$\\2. For each $(e_t, r, e') \in \mathcal{T}_{\mathcal{G}}$:\\ \quad \quad if $\cos(\mathbf{u}_t, \mathbf{v}_{e'}) \geq \tau$,\\ \quad \quad add $(e_t, r, e')$ to $\mathcal{T}_t$};

    \draw[arrow] (step) -- (llm);
    \draw[arrow] (trie) -- (lookup);
    \draw[arrow] (llm) -- (mask);
    \draw[arrow] (lookup) -- (mask);
    \draw[arrow] (mask) -- node[left, font=\footnotesize] {Sample $y_t$} (commit);

    \draw[arrow] (commit) -| node[pos=0.25, below, font=\footnotesize] {Trigger Expansion} (expand);
    \draw[arrow] (expand) |- node[pos=0.75, above, font=\footnotesize] {Update to $\mathcal{T}_t$} (trie);
  \end{tikzpicture}
  \caption[Step-Wise Dynamic Expansion Workflow (DCA‑Trie v2)]{Step‑Wise Dynamic Expansion Workflow for DCA‑Trie v2.}
  \label{fig:v2}
\end{figure}

\subsection{Algorithm}

\begin{figure}[H]
  \centering
  \begin{algorithm}[H]
    \caption{DCA‑Trie v2: Step‑Wise Semantic Expansion}
    \label{alg:v2}
    \begin{algorithmic}[1]
      \Require Knowledge graph $\mathcal{G}$; question entities $\mathcal{E}_q$;
      question $q$; admission threshold $\tau$; sentence encoder $E$;
      LLM with vocabulary $\mathcal{V}$
      \Ensure Generated reasoning chain $y_1 \ldots y_T$

      \State \textbf{Initialize trie} with entity nodes for $\mathcal{E}_q$:
      $\mathcal{T}_0 \gets \{e_q : e_q \in \mathcal{E}_q\}$

      \State \textbf{Precompute query embedding:}
      $\mathbf{u}_0 \gets E(\text{query}(q, \varepsilon))$

      \For{each step $t = 1$ \textbf{to} $T$}
      \State $\mathcal{V}_t \gets \textsc{TrieLookup}(\mathcal{T}_{t-1}, y_{<t})$
      \Comment{Valid tokens from current trie}
      \State $\mathbf{m}_t[v] \gets \mathbf{1}[v \in \mathcal{V}_t]$ for all $v$
      \State $\tilde{\boldsymbol{\alpha}}_t \gets \mathbf{m}_t \odot
        \textsc{LLMLogits}(y_{<t}, x)$
      \State $y_t \gets \text{beam-search-step}(\tilde{\boldsymbol{\alpha}}_t)$

      \If{$y_t$ is an \textit{entity} token $e_t$}
      \Comment{Expansion triggered at entity commits}
      \State \textbf{Update query embedding:}
      $\mathbf{u}_t \gets E(\text{query}(q, y_{\leq t}))$
      \State \textbf{Expand trie:}
      \For{each $(e_t, r, e') \in \mathcal{T}_{\mathcal{G}}$}
      \State $\mathbf{v}_{e'} \gets E(\text{str}((e_t, r, e')))$
      \Comment{Cache if previously computed}
      \State $s \gets \cos(\mathbf{u}_t, \mathbf{v}_{e'})$
      \If{$s \geq \tau$}
      \State Add $(e_t, r, e')$ to $\mathcal{T}_t$
      \EndIf
      \EndFor
      \Else
      \State $\mathcal{T}_t \gets \mathcal{T}_{t-1}$
      \Comment{No expansion at relation token steps}
      \EndIf
      \EndFor
      \State \Return $y_1 \ldots y_T$
    \end{algorithmic}
  \end{algorithm}
  \caption[Algorithm for DCA‑Trie v2]{DCA‑Trie v2 algorithm.}
  \label{fig:alg:v2}
\end{figure}

\subsection{Algorithm Description}

Unlike the static approach in v1, Algorithm~\ref{alg:v2} combines generation and semantic filtering. To manage how the search space changes, the process follows these steps:

\begin{enumerate}
  \item \textbf{Initialize a Minimal State:} The system starts with a basic trie that includes only the root question entities and a baseline query embedding.
  \item \textbf{Generate under Current Constraints:} At each step $t$, the LLM produces a new token that is strictly limited by the current physical boundaries of the trie.
  \item \textbf{Check for Entity Commitment:} The algorithm branches based on what the LLM just produced. If it generates an intermediate relation token, the trie stays the same and decoding continues. If the LLM finishes generating an entity token, it triggers an expansion phase.
  \item \textbf{Update the Context:} When an entity is committed, the system recalculates the query embedding. This new embedding merges the original question with the specific reasoning path the LLM just created, making the context very specific to the current state.
  \item \textbf{Dynamically Expand:} The system pulls all structurally valid neighbors of the newly committed entity from the master knowledge graph. It scores these candidates against the updated context embedding and discards the irrelevant ones. The remaining candidates are then added to the trie to enable valid future generation.
\end{enumerate}

This means we have a highly adaptable oracle. Instead of forcing the LLM to pick from a large, pre-calculated network of paths, the system builds the network just one step ahead of the model's current reasoning state, following its path while keeping the graph valid.

\subsection{Complexity Analysis}

This section presents a complexity analysis of DCA‑Trie v2 at the implementation level under the same assumptions. In v2, the expansion cost only happens when an entity token is committed, with a complexity of $O(|\mathcal{N}_{e_t}| \cdot d_E)$, where $|\mathcal{N}_{e_t}|$ is the neighborhood size of $e_t$. In practice, semantic filtering reduces this to about $O(|\mathcal{N}_{e_t}^{\text{filtered}}| \cdot d_E)$ with $|\mathcal{N}_{e_t}^{\text{filtered}}| \ll d$ for most nodes.

Relation-token steps do not trigger expansion, so their cost stays similar to that of standard constrained decoding. Caching further lowers overhead by preventing the repeated encoding of candidates that have already been seen.

\subsection{Faithfulness Guarantee}

As in v1, v2 maintains structural faithfulness by design. Every expansion candidate must meet the requirement $(e_t, r, e') \in \mathcal{T}_{\mathcal{G}}$ in Equation~\eqref{eq:v2expansion}. Semantic scoring only prunes this structurally valid set; it does not introduce new graph transitions. Therefore, all accepted tokens remain graph-based, and Condition~F holds true under proper token-mask construction and tokenizer alignment.

% -----------------------------------------------------------------------------
\section{Baseline Systems}
\label{sec:ch3baseline}
% -----------------------------------------------------------------------------

Four system configurations are evaluated to assess the proposed methodology:

\textbf{CoT Baseline:} The same Llama‑3.1‑8B model \citep{dubey2024llama3} used in GCR, run without any KG constraint oracle and prompted with standard chain-of-thought reasoning. This provides the unconstrained reference point.

\textbf{GCR:} The original Graph‑Constrained Reasoning framework \citep{luo2025-graph-constrained-reasoning} with static KG‑Trie constraints $\mathcal{W}_{\text{val}}(t)=f(\mathcal{G},\mathcal{E}_q)$. This is the primary constrained baseline. Reproduction targets WebQSP Hits@1 within 1--2\% of the reported 92.6\%.

\textbf{DCA‑Trie v1:} The static semantic filtering variant described in Section~\ref{sec:ch3v1}, implemented as a preprocessing step in the GCR pipeline with validation-calibrated threshold $\tau$.

\textbf{DCA‑Trie v2:} The dynamic step-wise expansion variant described in Section~\ref{sec:ch3v2}, integrated into the decoding process.

To ensure fair comparison, all settings use the same Llama‑3.1‑8B backbone, the same entity-linking pipeline for $\mathcal{E}_q$, and matched decoding settings (beam width $k=5$ for constrained systems). This isolates differences to the constraint-oracle design rather than model or preprocessing variation.

% -----------------------------------------------------------------------------
\section{Evaluation Protocol}
\label{sec:ch3eval}
% -----------------------------------------------------------------------------

\subsection{Datasets}

Evaluation is conducted on two standard Freebase-based KGQA datasets \citep{bollacker2008freebase}:

\textbf{WebQSP} \citep{yih2016websqp}: 4,737 questions, mostly requiring 1--2 hops. The standard test set is used for evaluation. A separate disjoint set of 100 questions is used only for threshold calibration.

\textbf{ComplexWebQuestions (CWQ)} \citep{talmor2018cwq}: 34,689 questions with higher compositional complexity and hop depths up to 4. This dataset is used in full for evaluation with no overlap with the calibration split.

\subsection{Evaluation Metrics}

Five metrics are used across both datasets:

\begin{enumerate}
  \item \textbf{Hits@1:} The proportion of examples where the top predicted answer matches the gold entity.

  \item \textbf{F1:} Entity-level overlap score for settings with multiple acceptable answers, following common KGQA evaluation practice \citep{lan2022survey}.

  \item \textbf{Structural Faithfulness Rate:} The fraction of generated paths where all triplets are valid in the underlying KG.

  \item \textbf{Semantic Irrelevance Ratio (SIR):} Computed using Equations~\eqref{eq:sir_step} and \eqref{eq:sir_aggregate} with a fixed reference threshold $\tau_{\text{ref}}=0.3$.

  \item \textbf{Average Trie Size Per Step:} The mean size of the valid token set $|\mathcal{V}_t|$ across decoding steps, used as a direct efficiency indicator of constraint selectivity.
\end{enumerate}

\subsection{Hop‑Depth Stratification}

All major metrics (except binary structural faithfulness) are also reported by hop depth (1, 2, 3, and 4), consistent with prior KGQA analysis \citep{lan2022survey}. This is important because permissiveness effects usually increase with hop depth; reporting only aggregate results can hide this behavior.

\subsection{Experimental Configuration}

All experiments are run on a single NVIDIA A100 (40GB). Llama‑3.1‑8B inference uses 16-bit precision through HuggingFace Transformers \citep{wolf2020transformers}. Constrained systems use beam width $k=5$, while the unconstrained CoT baseline uses $k=1$ to match the original GCR setup. The \texttt{all-MiniLM-L6-v2} encoder runs through `sentence-transformers` \citep{reimers2019sbert} on CPU to avoid GPU contention with LLM decoding. Maximum generation length is set to $3L+2$, with $L=2$ for WebQSP and $L=4$ for CWQ.

% -----------------------------------------------------------------------------
\section{Scope Boundaries and Anticipated Limitations}
\label{sec:ch3limitations}
% -----------------------------------------------------------------------------

This study has clear scope boundaries that frame the interpretation of the Chapter~4 results:

\textbf{Single KG and domain:} Experiments are limited to Freebase-based KGQA. Generalization to other KGs (e.g., Wikidata and ConceptNet) remains future work.

\textbf{No model fine-tuning:} The Llama‑3.1‑8B backbone is not fine-tuned. Observed improvements reflect changes in constraint-oracle design rather than adaptation of parameters.

\textbf{Empirical faithfulness evidence:} Condition~F is assessed empirically by checking generated paths against KG triplets, following established practice in prior work \citep{luo2025-graph-constrained-reasoning, li2024-decoding-on-graphs}. This evidence is interpreted under the same implementation assumptions regarding valid-token masking and tokenizer alignment.

\textbf{Threshold as a hyperparameter:} The selected $\tau$ is tuned on a validation subset and may not transfer well across datasets.

\textbf{False negative risk:} Even under Condition~P, semantic filtering can remove valid paths, particularly when lexical form and graph semantics diverge. This trade-off is explicitly analyzed in Chapter~4.

\textbf{Latency:} Semantic scoring adds extra compute overhead compared to GCR. Wall-clock latency is reported, but production deployment optimization is outside the scope of this thesis.