% =============================================================================
% CHAPTER 1 - INTRODUCTION
% =============================================================================

\chapter{INTRODUCTION}

\ifpdf
  \graphicspath{{introduction/figures/PNG}{introduction/figures/PDF}{introduction/figures}}
\else
  \graphicspath{{introduction/figures/EPS}{introduction/figures}}
\fi

\section{Background}

Large language models (LLMs) have emerged as one of the most consequential advances in artificial intelligence, demonstrating capabilities across language understanding, generation, and multistep reasoning that were considered beyond reach only a decade ago \citep{brown2020gpt3,openai2024gpt4technicalreport}. Models such as the GPT series \citep{brown2020gpt3,openai2024gpt4technicalreport} and the Llama family \citep{touvron2023llama,dubey2024llama} achieve strong performance on a broad spectrum of language tasks by learning statistical regularities over vast corpora of text. The scale and breadth of this training enables these models to exhibit apparent knowledge across diverse domains, from history and science to legal reasoning and medical information, without any explicit representation of factual knowledge in their architecture.

Despite these achievements, a fundamental and persistent failure mode undermines the reliability of LLMs in knowledge-intensive settings: hallucination. \citet{ji2023survey} define hallucination as the generation of content that is fluent and coherent but factually unsupported or contradicted by verifiable sources. \citet{huang2023survey} provide a comprehensive taxonomy distinguishing intrinsic hallucination, where the output contradicts a provided source, from extrinsic hallucination, where the output cannot be verified against any source. Critically, hallucination is not a rare edge case. \citet{luo2025-graph-constrained-reasoning} report that state-of-the-art retrieval-augmented approaches produce hallucinated reasoning steps in approximately one-third of cases even when granted access to a supporting knowledge graph. This failure rate renders unconstrained LLMs unsuitable for high-stakes knowledge-intensive tasks without additional mechanisms to enforce factual grounding. The structural inevitability of hallucination has been formally established in recent theoretical work. \citet{xu2025hallucinationinevitableinnatelimitation} prove that for any computable ground truth function, no LLM training procedure can achieve zero hallucination probability across all inputs--hallucination is an inherent property of computable learning systems rather than a transient training artifact. This "inevitability" result underscores that mitigation strategies relying solely on improved training, scaling, or prompting cannot fully eliminate hallucination; external structural constraints that ground generation in verifiable sources are theoretically necessary for guaranteed faithfulness. Knowledge graph-constrained decoding provides exactly such an external mechanism, making hallucination mathematically impossible rather than merely improbable.

The structural cause of hallucination is well understood. Autoregressive LLMs generate text by conditioning each token on all previously generated tokens, producing a probability distribution over the vocabulary at each step: $P(y_t \mid y_{<t}, x)$, where $x$ is the input and $y_{<t}$ is the partially generated sequence \citep{brown2020gpt3}. This process is effective at producing locally coherent text but provides no mechanism for grounding each generated claim in a verifiable external source. An error introduced at step $t$ propagates forward as the conditioning context for all subsequent tokens, producing confident and grammatically well-formed outputs whose factual incorrectness has no surface signal \citep{huang2023survey,ji2023survey}.

Knowledge graph question answering (KGQA) represents one of the most demanding contexts in which this failure mode manifests. A knowledge graph (KG) is a structured, human-curated database of verified facts encoded as (entity, relation, entity) triplets, for instance, (Marie Curie, award received, Nobel Prize in Physics) or (Paris, capital of, France), forming a large-scale graph over which multi-hop reasoning paths can be traced \citep{bollacker2008freebase}. The Freebase knowledge graph, which provides the underlying resource for the standard KGQA benchmarks WebQSP \citep{yih2016websqp} and ComplexWebQuestions \citep{talmor2018cwq}, encodes tens of millions of such triplets covering a wide range of domains and relation types. Answering a multi-hop question such as `What is the nationality of the director of the film in which the Blue Hawaii actor starred?' requires traversing a chain of three or more KG triplets in the correct order, a task for which an unconstrained LLM, relying entirely on parametric memory, may produce a plausible-sounding but factually incorrect chain that is indistinguishable in fluency from a correct one \citep{lan2022survey,he2021improving}.

The dominant paradigm for addressing this problem is retrieval-augmented generation (RAG), in which relevant passages or subgraphs are retrieved from an external source and provided to the LLM as additional context \citep{lewis2020rag,izacard2021leveraging}. While RAG reduces hallucination by providing relevant source material, it does not prevent the LLM from ignoring that material or generating claims that go beyond it. The LLM remains free to produce any token at any step; the retrieved context is a soft influence on its output, not a hard constraint on what it may generate \citep{asai2024selfrag,agrawal2024mindfulrag}.

KG-constrained decoding addresses this limitation directly. Rather than providing the KG as a soft contextual influence, constrained decoding frameworks interpose a constraint oracle between the LLM's logit distribution and the token selection mechanism. At each generation step, the oracle computes a binary mask over the vocabulary, setting to zero the logit of any token that would not correspond to a valid continuation of a KG path. The masking formulation, introduced by \citet{geng2023grammarconstrained} in the context of grammar-constrained decoding, takes the form $\tilde{\alpha}_t = m_t \odot \alpha_t$, where $\alpha_t$ is the raw logit vector, $m_t \in \{0,1\}^{\lvert V \rvert}$ is the validity mask, and $\odot$ denotes element-wise product. The generation process thus selects from a restricted set of valid tokens at every step, making structural hallucination, the generation of facts that do not exist in the KG, mathematically impossible rather than merely improbable \citep{willard2023outlines,geng2023grammarconstrained}.

The leading implementation of this paradigm is Graph-Constrained Reasoning (GCR) \citep{luo2025-graph-constrained-reasoning}, which achieves 92.6 Hits@1 on WebQSP and 75.8 Hits@1 on CWQ with verified 100\% structural faithfulness. GCR constructs a KG-Trie, a prefix tree \citep{fredkin1960trie} over all KG paths within $L$ hops of the question entities, obtained by breadth-first search, and uses it as the constraint oracle during beam-search decoding. Because every token in the generated reasoning chain must correspond to a valid prefix in the trie, the output is guaranteed to consist entirely of verified KG facts. This represents a qualitative improvement over both unconstrained LLMs and soft RAG approaches for the specific property of structural faithfulness.

However, a significant limitation remains. The KG-Trie in GCR is computed once, before generation begins, from the graph structure and the question entity set alone. It carries no information about the question semantics, the direction in which reasoning is proceeding, or which entities and relations have already been committed to in the partial generation $y_{<t}$. Formally, the valid token set $W_{\mathrm{val}}$ is a function $f(G, E_q)$, independent of $q$ and $y_{<t}$, and remains identical across every generation step. For a question about a specific narrow fact, this trie may contain hundreds or thousands of structurally valid but semantically irrelevant paths, forcing the LLM to discriminate among them using its own parametric knowledge: precisely the hallucination-prone mechanism that constrained decoding was designed to suppress \citep{li2024-decoding-on-graphs,luo2025-graph-constrained-reasoning}.

Decoding on Graphs (DoG)~\citep{li2024-decoding-on-graphs} introduces dynamic step-wise expansion, updating the constraint trie after each committed entity to include KG neighbors of the current node. This topological dynamism ensures the constraint set reflects the reasoning direction taken, addressing GCR's structural rigidity. DoG also employs question semantics, but crucially, only for \emph{ranking} candidate paths during beam search--the valid token set itself remains determined solely by graph connectivity, admitting every neighbor regardless of relevance to the question or partial generation. The permissiveness problem therefore persists: thousands of structurally valid but semantically irrelevant paths remain in the valid set at each step, forcing the LLM to discriminate among them using parametric knowledge. Furthermore, DoG sacrifices the computational efficiency of pre-computed tries, rebuilding the constraint structure at each decoding step. These limitations motivate the DCA-Trie framework, which applies semantic filtering directly to the constraint oracle, ensuring that the valid token set at each step is both structurally faithful and semantically relevant.

\section{Problem Statement}

Existing KG-constrained decoding frameworks, including GCR \citep{luo2025-graph-constrained-reasoning} and DoG \citep{li2024-decoding-on-graphs}, construct their constraint oracles as functions $f(G, E_q)$ of the knowledge graph and question entities, evaluated prior to or independently of the sequential generation process. This design has been shown to guarantee structural faithfulness (every generated token corresponds to a verified KG triplet), but provides no mechanism for adapting the constraint to the semantic content of the question or the direction taken by the partial generation \citep{geng2023grammarconstrained,willard2023outlines}.

The consequence is a fundamental conflation of two distinct requirements. Structural validity requires that every generated token correspond to a valid KG path prefix, a requirement all existing frameworks satisfy. Contextual relevance requires that the constraint set reflect the semantic focus of the reasoning at the current step, a requirement no existing framework satisfies. As a result, existing constraint oracles are simultaneously too broad, admitting structurally valid but semantically irrelevant paths that force the model to rely on parametric discrimination, and too rigid, failing to narrow in response to the evolving reasoning direction as generation progresses \citep{li2024-decoding-on-graphs,luo2025-graph-constrained-reasoning}.

This project addresses the following specific problem: current KG-constrained decoding frameworks do not condition the valid token constraint on the input question semantics or the partial generation, leaving the LLM responsible for semantic discrimination among a large set of valid-but-irrelevant paths at every generation step. This residual reliance on parametric knowledge partially re-introduces the hallucination risk that constrained decoding was designed to eliminate, and is most pronounced in multi-hop settings where the number of valid paths grows exponentially with hop depth \citep{lan2022survey,he2021improving,talmor2018cwq}.

Very recent work has independently addressed the efficiency-faithfulness trade-off in KGQA. RouterKGQA \citep{yuan2026routerkgqa} proposes routing between specialized and general models with semantic filtering at the answer level, achieving strong performance with minimal inference cost. While RouterKGQA and DCA-Trie both employ semantic relevance scoring, they operate at fundamentally different architectural stages: RouterKGQA filters \emph{answers} after generation, whereas DCA-Trie filters \emph{tokens} during generation. This distinction is critical: DCA-Trie's constraint oracle ensures the LLM never sees semantically irrelevant tokens, reducing cognitive load and enabling deterministic faithfulness guarantees that post-hoc filtering cannot provide. Furthermore, DCA-Trie introduces the Semantic Irrelevance Ratio (SIR; Chapter~\ref{Chapter3}), the first metric to quantify constraint permissiveness independently of downstream accuracy--an evaluation dimension not addressed by routing-based approaches.

To address these limitations, this project proposes DCA-Trie, upgrading the constraint oracle from $W_{\mathrm{val}}(t) = f(G, E_q)$ to $W_{\mathrm{val}}(t) = f(G, E_q, q, y_{<t})$. The upgraded framework must simultaneously satisfy three conditions: (i) structural faithfulness is preserved, every token admitted at each step corresponds to a valid KG path, as verified by exhaustive triplet checking against Freebase; (ii) semantic irrelevance is reduced, the proportion of the valid set consisting of paths irrelevant to the question decreases relative to GCR and DoG, as measured by the Semantic Irrelevance Ratio (SIR) introduced in this work; and (iii) answer recall is maintained: no gold answer path is erroneously excluded by the relevance filter, with false negative rate held below 5\% on the WebQSP validation set.

\section{Aim and Objectives}

This project aims to design, implement, and evaluate a Dynamic Context-Aware Trie (DCA-Trie) framework for knowledge graph-constrained large language model generation, in which the valid token constraint at each decoding step is conditioned jointly on the knowledge graph structure, the input question semantics, and the partially generated reasoning chain, achieving improved answer accuracy on multi-hop KGQA benchmarks while maintaining 100\% structural faithfulness to the underlying knowledge graph.

To achieve this aim, the following objectives have been set:

\begin{enumerate}[label=\roman*.]
  \item To characterize the permissiveness problem in existing KG-constrained decoding frameworks by defining and measuring the Semantic Irrelevance Ratio (SIR) of GCR's static KG-Tries on the WebQSP and CWQ benchmarks, stratified by hop depth.
  \item To design a semantic relevance scoring mechanism using pretrained sentence transformer embeddings \citep{reimers2019sbert} that can filter candidate KG paths by joint relevance to the input question and partial generation at inference time, without task-specific fine-tuning.
  \item To implement a static semantic filtering variant (DCA-Trie v1) that applies the relevance scorer at KG-Trie construction time, reducing trie size before generation begins while maintaining a false negative rate below 5\% on gold answer paths on the WebQSP validation set.
  \item To implement a step-wise dynamic expansion variant (DCA-Trie v2) in which the trie is updated after each committed triplet using the relevance scorer conditioned jointly on the question and the partial generation $y_{<t}$, directly operationalizing the $W_{\mathrm{val}}(t) = f(G, E_q, q, y_{<t})$ formulation.
  \item To evaluate DCA-Trie v1 and DCA-Trie v2 against the GCR baseline and an unconstrained chain-of-thought (CoT) baseline on WebQSP and CWQ, measuring Hits@1, F1, structural faithfulness, SIR, and average trie size, with results stratified by question hop depth.
  \item To demonstrate the practical applicability of DCA-Trie by building an experimental interactive prototype system (web-based or command-line QA interface) for research demonstration, allowing users to input complex natural language questions and receive factually grounded reasoning chains and answers, thereby showcasing the framework's real-world value beyond benchmark metrics.
\end{enumerate}

\section{Tools and Facilities Used}

\subsection{Software Tools}

Table \ref{tab:implemented-software-tools} presents an evaluation of the software tools implemented in this project, their role within the DCA-Trie framework, and the rationale for their selection.

\begin{table}[H]
  \centering
  \small
  \setlength{\tabcolsep}{4pt}
  \caption{Implemented Software Tools Evaluation}
  \label{tab:implemented-software-tools}
  \begin{tabularx}{\textwidth}{|>{\raggedright\arraybackslash}p{0.20\textwidth}|>{\raggedright\arraybackslash}p{0.30\textwidth}|>{\raggedright\arraybackslash}X|}
    \hline
    Tool                                     & Use in Project                                                                                                    & Rationale                                                                                                                       \\
    \hline
    Python / HuggingFace Transformers        & Model inference pipeline; loading and running GCR's Llama-3.1-8B checkpoint; integrating the relevance scorer     & Provides reliable model loading, tokenization, and generation utilities, and enables direct use of the released GCR checkpoint. \\
    \hline
    GCR Framework                            & Primary baseline; KG-Trie construction, graph-constrained decoding, and result reproduction on WebQSP and CWQ     & Provides a strong baseline and supports the trie-based constrained decoding mechanism that DCA-Trie extends.                    \\
    \hline
    Sentence-transformers / all-MiniLM-L6-v2 & Semantic relevance scoring; computing cosine similarity between question-context and candidate KG path embeddings & Provides fast semantic embeddings without extra fine-tuning and supports efficient iterative decoding.                          \\
    \hline
    PyTorch                                  & Deep learning framework underpinning all model inference and custom decoding logic                                & Supports implementation and testing of custom decoding logic and integrates smoothly with HuggingFace.                          \\
    \hline
    Freebase / WebQSP and CWQ Datasets       & Knowledge graph source and KGQA benchmarks for evaluation                                                         & Provides standard datasets for this research area and enables direct comparison with prior work.                                \\
    \hline
    Google Colab Pro (A100, 40 GB)           & GPU environment for all inference experiments; Llama-3.1-8B constrained decoding with beam search                 & Provides enough GPU memory for stable experiments and removes the need for local high-end hardware.                             \\
    \hline
    Visual Studio Code + Git / GitHub        & Development environment and version control for all four team members                                             & Supports day-to-day coding, team collaboration, version tracking, and reproducibility throughout the project.                   \\
    \hline
  \end{tabularx}
\end{table}

\subsection{Facilities}

The following facilities were utilized in the execution of this project:

\begin{enumerate}[label=\roman*.]
  \item Google Colab Pro (A100, 40 GB VRAM) for all model inference and constrained decoding experiments.
  \item University library and online academic databases (ACM Digital Library, arXiv, Google Scholar, ACL Anthology) for literature review and citation verification.
  \item University internet infrastructure for cloud connectivity, dataset access, and retrieval of model checkpoints from the HuggingFace model hub.
  \item Shared team repository hosted on GitHub for version control, experiment logging, and reproducibility of all reported results.
\end{enumerate}

\section{Methods Used}

The project employs the following methods:

\begin{enumerate}[label=\roman*.]
  \item Systematic review and critical analysis of the academic literature on constrained decoding, knowledge graph question answering, hallucination in LLMs, and retrieval-augmented generation.
  \item Formal gap analysis of the four foundational papers, GCR \citep{luo2025-graph-constrained-reasoning}, DoG \citep{li2024-decoding-on-graphs}, GCD \citep{geng2023grammarconstrained}, and Outlines \citep{willard2023outlines}, to identify the specific limitations that motivate the DCA-Trie design, including the permissiveness problem and the absence of generation-state feedback in the constraint oracle.
  \item Formal specification of the DCA-Trie framework, including definition of the Semantic Irrelevance Ratio metric, the relevance scoring mechanism using sentence transformer embeddings \citep{reimers2019sbert}, the static filtering procedure for v1, and the step-wise dynamic expansion rule for v2.
  \item Reproduction of the GCR baseline on the WebQSP benchmark using GCR's publicly released Llama-3.1-8B checkpoint to establish a verified starting point within 1--2\% of published Hits@1.
  \item Implementation of DCA-Trie v1 and v2 as extensions to GCR's inference pipeline, with relevance threshold selected on a 100-question held-out validation subset of WebQSP to minimize false negative rate on gold answer paths.
  \item Quantitative evaluation of all system configurations on WebQSP and CWQ using Hits@1, F1, faithfulness rate, SIR, and average trie size per step, stratified by question hop depth following the evaluation protocol of \citet{lan2022survey}.
  \item Ablation studies examining the sensitivity of DCA-Trie's performance to the relevance threshold $\tau$, the sentence encoder architecture, and beam size $k$, isolating the contribution of each mechanism to the overall result.
  \item Qualitative error analysis identifying failure cases in which DCA-Trie's semantic filter erroneously excludes a gold answer path, supporting honest characterization of the framework's limitations and providing concrete directions for future work.
\end{enumerate}

\section{Scope of Work}

This project designs and evaluates a Dynamic Context-Aware Trie framework for KG-constrained LLM decoding, with the specific aim of addressing the permissiveness problem identified in GCR and DoG. The framework is developed and evaluated in the context of multi-hop KGQA over Freebase-grounded benchmarks (WebQSP and CWQ), which are the standard evaluation settings in the KG-constrained decoding literature. All experiments use GCR's publicly released Llama-3.1-8B checkpoint; no new model training or fine-tuning is performed. The semantic relevance scorer uses the all-MiniLM-L6-v2 sentence transformer without domain-specific adaptation.
We emphasize that all experiments use GCR's publicly released Llama-3.1-8B checkpoint \citep{luo2025-graph-constrained-reasoning}, which includes their fine-tuned KG-specialized parameters. No additional model training or fine-tuning is performed in this work; our contribution is strictly the constraint oracle architecture and semantic filtering mechanism.

The scope is explicitly bounded as follows. Generalization of DCA-Trie to knowledge graphs other than Freebase, or to non-KGQA tasks such as entity linking or structured information extraction, is identified as future work and is not evaluated in this dissertation. Formal proofs of the faithfulness guarantee under dynamic pruning are beyond scope; empirical measurement of the gold answer false negative rate is provided as evidence. The confidence-conditioned constraint tightening mechanism, in which constraint granularity is modulated by the LLM's token-level entropy, is described as a theoretical extension but is not implemented. Production-scale deployment, latency optimization for online serving, and integration with proprietary enterprise knowledge graphs are engineering extensions beyond the research scope of this project.

The primary contribution of this work is the DCA-Trie framework and the Semantic Irrelevance Ratio metric. Secondary contributions are the empirical characterization of constraint permissiveness in existing frameworks and the ablation analysis isolating the contribution of semantic filtering from dynamic expansion. All code, experimental configurations, and the curated evaluation datasets will be made publicly available to support reproducibility and future extension.

\section{Organization of Work}

Chapter 1 introduces the study by identifying the core problem, stating the project objectives, and defining the intended DCA-Trie solution, thereby establishing the motivation, scope, and overall direction of the research. Chapter 2 presents the literature review by examining prior work on hallucination in LLMs, KGQA, constrained decoding, and related frameworks, and it identifies key limitations in existing approaches to justify the need for a context-aware constraint mechanism. Chapter 3 combines problem formulation and methodology, defining the upgraded constraint setup, research questions, evaluation criteria, and implementation strategy for DCA-Trie v1 and v2 within the GCR pipeline \citep{luo2025-graph-constrained-reasoning}. Chapter 4 presents results and discussion, reporting quantitative evaluation across WebQSP and CWQ, analysing faithfulness and semantic irrelevance trends, and interpreting the findings through ablation and error analysis. Chapter 5 concludes the dissertation by summarizing the main contributions, highlighting practical implications, and presenting recommendations for future work.
