% =============================================================================
% CHAPTER 2: LITERATURE REVIEW
% =============================================================================

\chapter{Literature Review}

% -----------------------------------------------------------------------------
\section{Introduction}

Large language models (LLMs) have fundamentally transformed natural language processing, demonstrating remarkable capabilities across a wide range of tasks. However, their reliance on autoregressive generation introduces a critical vulnerability: the propensity to confidently generate fluent but factually incorrect information, known as hallucination~\citep{ji2023survey}. In domains requiring precise, verifiable reasoning, such as knowledge-intensive question answering, this lack of factual reliability remains a significant barrier. To address this, the field has increasingly looked toward grounding language models in structured, external facts using knowledge graphs. Yet, ensuring that a model remains strictly faithful to these graphs during generation, without sacrificing fluency, remains an open and complex challenge.

The purpose of this chapter is to survey the existing literature that contextualizes and motivates the research problem this dissertation addresses: how to structurally guarantee factual faithfulness in multi-hop language model reasoning. Rather than describing the proposed system itself, this review maps the intersecting research landscapes of language model hallucination, multi-step reasoning, and structured knowledge grounding. By critically examining current mitigation strategies, ranging from advanced prompting to retrieval-augmented generation and constrained decoding frameworks, this chapter establishes the fundamental limitations of approaches that rely solely on soft grounding or static constraints, and identifies the specific gaps that remain open in the existing literature.

The review is structured as follows. Section~\ref{sec:llms} covers the fundamentals of Large Language Models. Section~\ref{sec:hallucination} establishes hallucination in large language models as a structural problem inherent to autoregressive generation. Section~\ref{sec:cot} examines chain-of-thought reasoning and its extensions, showing why improved prompting strategies cannot provide structural faithfulness guarantees. Section~\ref{sec:kgqa} introduces knowledge graphs and knowledge graph question answering, including the benchmarks and the complexity properties that make multi-hop KGQA challenging. Section~\ref{sec:constrained} traces the development of constrained decoding from foundational logit masking through grammar constraints to knowledge graph-specific frameworks. Section~\ref{sec:rag} reviews retrieval-augmented generation and establishes why soft grounding approaches cannot achieve structural faithfulness. Section~\ref{sec:sbert} reviews semantic similarity scoring with sentence transformers. Section~\ref{sec:related} surveys related works organized by approach and critiques each cluster. Section~\ref{sec:synthesis} synthesizes the three gaps that remain unresolved across the literature.

% -----------------------------------------------------------------------------
\section{Large Language Models}
\label{sec:llms}
% -----------------------------------------------------------------------------

To understand how knowledge graph constraints can be applied to language models, it is first necessary to understand the underlying mechanism of language model generation. Large language models (LLMs) are the foundational engines driving modern natural language processing. This section establishes the architectural and generative properties of LLMs, specifically detailing the autoregressive process that gives rise to both their fluent reasoning capabilities and their vulnerability to factual hallucination.

\subsection{Transformer Architecture}

Modern LLMs are built upon the Transformer architecture, introduced by \citet{vaswani2017attention}. At its core, the Transformer relies on a self-attention mechanism that allows the model to dynamically weigh the importance of different tokens in an input sequence regardless of their positional distance. This represented a fundamental shift from prior recurrent architectures, enabling highly parallelized training and the capture of long-range dependencies. While the original framework proposed an encoder-decoder structure designed for translation tasks, the field has largely converged on decoder-only variants for text generation. In a decoder-only model, self-attention is causally masked, meaning that when computing the representation for a given token, the model can only attend to previous tokens and not future ones. This strictly left-to-right processing naturally aligns with the task of predicting the next word in a sequence. The Llama-3.1-8B model~\citep{dubey2024llama3} is a prominent example of this decoder-only, causally-masked transformer design, and has been adopted as the backbone in recent KG-constrained decoding work~\citep{luo2024gcr}.

\subsection{Autoregressive Generation}

The text production process in these models is fundamentally autoregressive. Given a vocabulary $\mathcal{V}$, an input sequence $x$, and previously generated output tokens $y_{<t}$, the model computes a probability distribution over the entire vocabulary for the next token: $P(y_t | y_{<t}, x)$. Because the generation condition includes all previously generated tokens, every new word becomes part of the context for the next word. Decoding strategies, such as beam search and nucleus sampling~\citep{holtzman2020curious}, determine how a specific token is selected from this computed distribution. Crucially, because this distribution is always computed over the full vocabulary, any mathematically possible token has a strictly positive, non-zero probability of being generated at any step. This autoregressive property is the mechanism that produces highly fluent, coherent text; however, because each step conditions on prior outputs without an external verifier, it is also the direct structural cause of hallucination.



\subsection{Large-Scale Pretraining}

The transition from basic transformers to ``large'' language models is driven by scale in both parameter count and training data. Foundational models are pretrained on massive corpora of unlabelled text using a simple self-supervised objective, typically predicting the next token. The GPT series (\citealt{brown2020gpt3}; \citealt{achiam2023gpt4}) and the open-weight Llama series (\citealt{touvron2023llama}; \citealt{dubey2024llama3}) demonstrated that scaling this simple objective leads to the emergence of powerful downstream capabilities, including few-shot learning and broad world knowledge. During pretraining, the model internalises patterns, facts, and logical structures present in its training distribution. However, this parametric memory is fundamentally static and statistical. While scaling up models improves their fluency, syntax, and ability to recall commonly observed facts, it does not fix the underlying generation mechanism. The model still lacks a structural way to interact with dynamic external facts or guarantee that its statistical outputs map to verified truth in the real world.

\subsection{Instruction Following and Chain-of-Thought}

To align these pretrained base models with user intent, they undergo post-training processes such as instruction tuning and reinforcement learning from human feedback (RLHF). This alignment allows models to engage in dialogue and follow complex instructions. One of the most significant capabilities elicited from aligned LLMs is multi-step reasoning, typically prompted via Chain-of-Thought (CoT; \citealt{wei2022cot}). As surveyed by \citet{huang2023towards}, CoT prompting encourages the model to generate intermediate reasoning steps before arriving at a final answer, effectively breaking down complex problems. While CoT dramatically improves performance on logic and reasoning benchmarks, it suffers from the same autoregressive limitation: each intermediate reasoning step is generated using only the model's internal statistical distributions. Therefore, while CoT helps structure a language model's reasoning process, it cannot logically ground those outputs in verifiable sources. This persistent lack of factual grounding creates the necessity for more rigorous interventions, such as knowledge graph-constrained decoding.

% -----------------------------------------------------------------------------
\section{Hallucination in Large Language Models}
\label{sec:hallucination}
% -----------------------------------------------------------------------------

\subsection{Definition and Taxonomy}

Hallucination in natural language generation refers to the production of content that is fluent and internally coherent yet factually unsupported by or in outright contradiction with verifiable sources~\citep{ji2023survey}. \citet{huang2023hallucination} refine this definition into a two-part taxonomy. \textit{Intrinsic hallucination} occurs when the model output directly contradicts material present in the provided source context. \textit{Extrinsic hallucination} occurs when the output contains claims that cannot be verified against any external source, even if not directly contradicting a specific reference.

Both types are documented at non-trivial rates in deployed systems. \citet{ji2023survey} survey over 30 works across summarization, dialogue, and question answering, finding hallucination rates ranging from 15\% to over 60\% depending on task type and measurement methodology. In the specific context of knowledge graph question answering, \citet{luo2024gcr} report that retrieval-augmented reasoning approaches produce hallucinated reasoning steps in approximately one-third of cases even when provided direct access to a supporting knowledge graph, underscoring that the problem is not simply one of insufficient external knowledge.

\subsection{Structural Causes: The Autoregressive Generation Mechanism}

The root cause of hallucination lies in the autoregressive generation mechanism shared by all modern large language models. At each generation step, the model computes a probability distribution over its vocabulary conditioned on the input and all previously generated tokens. Three structural properties of this mechanism make hallucination an ever-present risk rather than an occasional failure.

First, the generation distribution is computed entirely from learned parameters, with no term that corresponds to a factual verification operation against an external source. The model has no mechanism to ``check'' a candidate token against a knowledge base before selecting it. Second, errors propagate forward: because each step is conditioned on all prior output, a hallucinated token at step $t$ becomes part of the conditioning context for step $t{+}1$, producing fluent and internally consistent continuations of an already-incorrect chain. Third, the distribution is defined over the full vocabulary at every step, meaning that any token has a strictly positive probability of selection at any step, regardless of factual correctness~\citep{brown2020gpt3,huang2023hallucination,ji2023survey}.

\subsection{The Scaling Paradox}

A common objection to the severity of the hallucination problem is that sufficiently large models should eventually achieve adequate factual reliability. The evidence does not support this view. \citet{lin2022truthfulqa} introduce TruthfulQA, a benchmark targeting questions that frequently elicit plausible but false responses. Their central finding is that larger models do not outperform smaller models on truthfulness and in certain categories exhibit \textit{inverse scaling}; in other words, larger models perform worse because they more effectively reproduce prevalent but incorrect beliefs from training data.

\citet{perez2022sycophancy} identify a compounding problem in models trained with reinforcement learning from human feedback: such models learn to generate confident-sounding responses that align with apparent user expectations rather than with verifiable facts, decoupling confidence from accuracy. \citet{kandpal2023large} show that while larger models memorize more training-data facts and improve on frequently-seen entity combinations, they do not improve on compositional questions requiring combinations of facts not seen during training. This precisely mirrors the structure of multi-hop knowledge graph questions. \citet{mallen2023not} confirm substantial hallucination rates for GPT-4 on questions involving tail entities, regardless of model scale. Together, these findings establish that hallucination is a structural feature of the autoregressive mechanism, not a scale problem with a straightforward remedy.

% -----------------------------------------------------------------------------
\section{Multi-Step Reasoning: Chain-of-Thought and Its Limits}
\label{sec:cot}
% -----------------------------------------------------------------------------

\subsection{Chain-of-Thought Prompting}

Chain-of-thought (CoT) prompting, introduced by \citet{wei2022cot}, is the dominant paradigm for eliciting multi-step reasoning from large language models. By including worked examples in the prompt that show intermediate reasoning steps, the model learns to decompose complex problems into a sequence of simpler sub-problems before producing a final answer. \citet{wei2022cot} demonstrate substantial accuracy improvements on arithmetic, commonsense, and multi-hop question answering benchmarks, with the capability emerging at approximately 100 billion parameters. Zero-shot chain-of-thought~\citep{kojima2022zero} achieves similar benefits without worked examples by adding the phrase ``Let's think step by step'' to the prompt, demonstrating that the underlying capability is latent in sufficiently large models and requires only elicitation.

The limitation of CoT is that each intermediate step is generated by the same autoregressive mechanism described in the previous section. An incorrect intermediate step propagates forward as conditioning context, and no mechanism verifies that any step corresponds to a fact in an external knowledge source. A CoT chain can be entirely fluent and internally coherent while containing one or more steps that are factually fabricated.

\subsection{Extensions: Self-Consistency and Tree-of-Thought}

Two major extensions of chain-of-thought address its reliability limitations without resolving its structural faithfulness problem. Self-consistency~\citep{wang2023selfconsistency} samples multiple independent CoT chains and selects the final answer by majority vote, marginalizing over individual chain errors. This approach substantially improves accuracy on benchmarks such as GSM8K and multi-hop QA by reducing sensitivity to individual hallucinated chains. However, it provides no guarantee that any single chain is factually grounded: a majority of independently hallucinated chains can still produce an incorrect majority-vote answer, particularly on deep multi-hop questions where error probabilities compound.

Tree-of-Thought~\citep{yao2023tree} frames generation as a search over a tree of partial reasoning sequences, using the language model as both a generator and an evaluator to prune unpromising branches. This improves performance on tasks requiring backtracking and exploration. The fundamental limitation remains: pruning decisions are made by the model itself, using its own parametric knowledge as the sole evaluator, without any external verification of the factual status of candidate reasoning steps.

\subsection{The Faithfulness Ceiling of Prompting-Based Approaches}

The common limitation of all prompting-based approaches, including few-shot, chain-of-thought, self-consistency, tree-of-thought, and their combinations, is that they modify the \textit{input} to the language model without modifying the \textit{generation mechanism}. Because decoding still occurs over the full vocabulary, every token retains a strictly positive probability at every step. True structural faithfulness, defined as the guarantee that every generated token corresponds to a verified external fact, requires that invalid tokens be assigned exactly zero probability. This can only be achieved by intervening directly in the decoding algorithm, which is the function of constrained decoding approaches~\citep{geng2023gcd, willard2023outlines}.

% -----------------------------------------------------------------------------
\section{Knowledge Graphs and Knowledge Graph Question Answering}
\label{sec:kgqa}
% -----------------------------------------------------------------------------

\subsection{Knowledge Graphs}

A knowledge graph is a structured database of verified facts, represented as a set of directed, labelled triplets $(e, r, e')$ where $e, e' \in \mathcal{E}$ are entities and $r \in \mathcal{R}$ is a relation type. Each triplet encodes a single human-verified fact: for example, (\textit{Marie Curie}, \textit{award\_received}, \textit{Nobel\_Prize\_Physics}) or (\textit{Blue\_Hawaii}, \textit{film.film.featured\_film\_location}, \textit{Hawaii}). Freebase~\citep{bollacker2008freebase}, the knowledge graph underlying the evaluation benchmarks used in this work, encodes tens of millions of such triplets across geography, history, entertainment, and science, and has been widely adopted as the standard resource for KGQA research~\citep{yih2016websqp, talmor2018cwq}.

The critical distinction between a knowledge graph and an unstructured text corpus is that every triplet in the KG has been explicitly verified and is machine-readable in a deterministic, unambiguous form. This property makes KGs suitable as sources for hard constraints on generation: a candidate token can be verified against the KG via an exact lookup, rather than via probabilistic semantic matching against raw text. A multi-hop reasoning path of length $L$ through the KG is a sequence $(e_0, r_1, e_1, \ldots, r_L, e_L)$ in which each consecutive triplet $(e_{i-1}, r_i, e_i)$ is a verified member of $\mathcal{T}$~\citep{lan2022kgqa}.

\subsection{Knowledge Graph Question Answering}

Knowledge graph question answering (KGQA) requires a system to identify the answer entity $e_L$ to a natural language question $q$ by reasoning through a chain of verified KG triplets. The task is formally defined as: given question $q$ with associated question entities $\mathcal{E}_q \subseteq \mathcal{E}$ identified by entity linking, identify the terminal entity of the correct reasoning path in $\mathcal{G}$~\citep{lan2022kgqa}.



Early KGQA approaches relied on either semantic parsing (converting questions to structured SPARQL queries) or information retrieval over KG subgraphs~\citep{lan2022kgqa}. More recent approaches use large language models as retrievers, agents iteratively querying the KG, or generators conditioned on retrieved subgraph context~\citep{jiang2023structgpt, jin2024graphcot, luo2024rog}. The shift toward LLM-based KGQA has improved fluency and coverage but reintroduced the hallucination risk that structured approaches avoided: LLM-generated reasoning chains are not guaranteed to correspond to verified KG paths.

\subsection{Evaluation Benchmarks}

Two benchmark datasets provide the evaluation setting for KGQA research. WebQSP~\citep{yih2016websqp} is derived from the WebQuestions dataset and comprises 4,737 questions grounded in Freebase with reasoning depths of one to two hops, annotated with SPARQL queries and evaluated using Hits@1 and F1. ComplexWebQuestions (CWQ;~\citealt{talmor2018cwq}) extends WebQSP by compositionally transforming questions through conjunction, composition, and comparative operations, yielding approximately 34,000 questions requiring up to four reasoning hops. WebQSP represents a regime where the set of valid paths is comparatively small; CWQ represents the regime where valid path sets grow large and the difficulty of identifying the correct path among structurally valid alternatives is highest.

\subsection{Multi-Hop Complexity}

The combinatorial difficulty of multi-hop KGQA increases sharply with hop depth. For an entity with average out-degree $d$ in the KG, the number of valid paths of length $L$ reachable from the question entities grows as $|\mathcal{E}_q| \times d^L$. For well-connected Freebase entities where $d \approx 20$ and a question with up to three question entities, a three-hop question can produce up to 24{,}000 structurally valid paths, of which exactly one leads to the correct answer~\citep{he2021multihop, lan2022kgqa}. This combinatorial explosion has concrete consequences for any approach that must discriminate among candidate paths: at each generation step, the model must select the correct next token from among a very large set of valid-but-incorrect alternatives, relying on its parametric knowledge to do so.

% -----------------------------------------------------------------------------
\section{Constrained Decoding for Faithful Text Generation}
\label{sec:constrained}
% -----------------------------------------------------------------------------

Constrained decoding is the paradigm in which generation is restricted to a predefined valid set at each step through direct intervention in the decoding mechanism. Unlike prompting, which modifies the input to the LLM, constrained decoding modifies the token selection process itself: invalid tokens are assigned zero probability before sampling, making their generation structurally impossible.

\subsection{Logit Masking}

The foundational mechanism of constrained decoding is logit masking, described formally in \citet{geng2023gcd} and \citet{willard2023outlines}. A constraint oracle computes a binary validity mask over the vocabulary at each step; the mask is applied element-wise to the raw logit vector before softmax normalization, setting invalid token logits to negative infinity and thereby assigning them zero probability. Token selection then operates exclusively over the valid subset. The central design question in any constrained decoding system is how the validity mask is computed: what information the oracle uses, and how it changes across generation steps.

\subsection{Grammar-Constrained Decoding}

\citet{geng2023gcd} introduce Grammar-Constrained Decoding (GCD), which constrains generation to strings conforming to a context-free grammar, using an Earley parser as the oracle. The valid next-token set at each step is the set of tokens that extend the current partial string as a valid grammar prefix. GCD additionally introduces input-dependent grammars, in which the grammar is parameterized by the input question, representing a partial step toward context-sensitive constraints. \citet{willard2023outlines} address the computational cost of constrained decoding with an efficient finite-state machine formulation. By precomputing an index from automaton states to valid vocabulary subsets, the valid token set can be retrieved in amortized constant time at each generation step, making constrained decoding practical at inference speed.

The shared limitation of these format-constraint approaches is that they enforce structural validity of the output format (such as grammatical correctness or schema conformance) without enforcing factual correctness. A grammatically valid output can contain entirely hallucinated facts. The transition from format constraints to factual constraints requires a constraint oracle that verifies tokens against a knowledge source rather than against a grammar.

\subsection{Knowledge Graph-Constrained Decoding}

The application of constrained decoding to knowledge graph faithfulness is a more recent development. The core idea is to use the KG itself as the constraint oracle: at each generation step, only tokens that correspond to valid continuations of a KG reasoning path are admitted. This approach guarantees structural faithfulness, meaning every generated token corresponds to a verified KG triplet, which makes hallucination of KG facts structurally impossible.

\textbf{Graph-Constrained Reasoning (GCR)}~\citep{luo2024gcr} is the state of the art in this line of work. GCR constructs a KG-Trie (a prefix tree over all KG paths within $L$ hops of the question entities, built by breadth-first search before generation begins) and uses it as the constraint oracle during beam-search decoding. The trie is queried at each generation step to return valid next tokens. GCR achieves 92.6 Hits@1 on WebQSP and 75.8 on CWQ with verified 100\% structural faithfulness using a Llama-3.1-8B backbone. A key property is that the trie is constructed once before generation and never modified: the valid token set is fixed regardless of what has been generated.

\textbf{Decoding on Graphs (DoG)}~\citep{li2024dog} introduces step-wise trie expansion as a partial remedy to GCR's static oracle. DoG defines \textit{well-formed chains} as sequences of KG triplets in which every triplet exists in the KG and every entity is connected to the preceding triplet or is a question entity. It enforces this definition through step-wise expansion: after committing to an entity, the constraint trie is extended with all KG neighbors of that entity. This makes the oracle topologically dynamic, growing as reasoning progresses through the graph. DoG achieves 90.99 Hits@1 on WebQSP and 76.08 on CWQ with 100\% structural faithfulness, and is training-free, requiring no fine-tuning. However, the expansion rule admits all KG neighbors of each visited entity regardless of their relevance to the question or the current reasoning direction; and per-step trie reconstruction sacrifices the efficiency of precomputed constraint structures.

\textbf{RouterKGQA}~\citep{yuan2026routerkgqa} takes a different approach, routing between a specialised KG-reasoning model and a general LLM based on question complexity. It applies semantic filtering using SBERT embeddings (specifically nomic-embed-text-v1) to select among candidate answers after generation, achieving competitive accuracy with reduced inference cost. RouterKGQA's semantic scoring operates at the answer selection level rather than at the token constraint level: paths are generated first and filtered retrospectively, rather than being constrained during generation.

\subsection{Beam Search and Trie-Based Constraints}

To understand how knowledge graph constraints are practically enforced during generation, it is essential to examine the two primary search algorithms and data structures involved: beam search and the prefix trie.

\textbf{Beam Search.} First introduced in the context of the HARPY speech recognition system~\citep{lowerre1976harpy}, beam search is a heuristic search algorithm used in natural language processing to decode sequences from probabilistic models. Unlike greedy decoding, which exclusively selects the single most probable token at each step and risks generating sub-optimal global sequences, beam search maintains a set of the $k$ most promising partial hypotheses (or ``beams") at any given time. At each generation step, the algorithm expands all $k$ current hypotheses by scoring their possible continuations, ranks the resulting sequences based on their cumulative log-probability, and prunes the search space back down to the top $k$ candidates~\citep{holtzman2020curious}. This approach mitigates the risk of early compounding errors while remaining computationally tractable, making it the dominant decoding strategy for tasks requiring structured and coherent output.

\textbf{The Prefix Trie.} A trie, originating from the concept of a ``trie memory" (retrieval) introduced by \citet{fredkin1960trie}, is a deterministic tree-based data structure used to store a dynamic set of sequences, typically strings. In a trie, no single node stores a complete string; instead, a node's position in the tree defines the exact prefix it represents. The branches extending from any given node represent the deterministic set of valid next characters or tokens. In the domain of constrained decoding, a pre-computed trie serves as a highly efficient constraint oracle. Given the current generated sequence prefix, traversing the trie to the corresponding node yields the exact subset of valid next tokens in amortised constant time, $O(1)$, relative to vocabulary size. This operation is critically more efficient than running continuous dynamic parser evaluations, making trie lookups the most scalable pattern for enforcing hard factual constraints at inference speed~\citep{willard2023outlines}.

\textbf{The KG-Trie.} For knowledge graph-constrained generation, architectures such as Graph-Constrained Reasoning (GCR) utilize a specialized adaptation known as the KG-Trie~\citep{luo2024gcr}. A KG-Trie serializes structurally verified multi-hop paths from the knowledge graph into string formats and maps them across the language model's precise token vocabulary. Rather than nodes representing single alphabetical characters, each edge in the KG-Trie represents a specific valid token in the LLM's vast vocabulary space.

When beam search is merged with a KG-Trie oracle (as visualized in Figure~\ref{fig:kg_trie_search}), the decoding process transforms into a strictly bounded traversal of the prefix tree. At each autoregressive step, the current beam hypotheses map to specific states in the KG-Trie. The language model calculates a probability distribution over its entire vocabulary, but a mathematical mask forces the probability of any candidate token not validly present as a child node in the trie to exactly zero ($-\infty$ prior to softmax normalization). The beam search then advances strictly down the permissible branches, functionally restricting the LLM's autoregressive generation to a set of pre-verified deterministic paths while still allowing the model to express internal preference (via log-probabilities) among those factually valid chains.

\begin{figure}[htbp]
    \centering
    \resizebox{0.75\textwidth}{!}{
        \begin{tikzpicture}[
            node distance=1.2cm and 1.0cm,
            trie/.style={circle, draw, minimum size=1.1cm, inner sep=1pt, align=center, font=\scriptsize\ttfamily},
            leaf/.style={circle, draw, fill=gray!20, minimum size=1.1cm, inner sep=1pt, align=center, font=\scriptsize\ttfamily},
            arrow/.style={-{Stealth[scale=1.0]}, thick},
            invalid/.style={circle, draw, dashed, text=gray, minimum size=1.1cm, inner sep=1pt, align=center, font=\scriptsize\ttfamily},
            invalidarrow/.style={-{Stealth[scale=1.0]}, dashed, draw=gray}
            ]

            % Root
            \node[trie, fill=blue!10] (root) {State\\$s_0$};

            % Level 1
            \node[trie, below left=1.2cm and 1.5cm of root] (t1) {$\langle$par$\rangle$};
            \node[trie, below right=1.2cm and 1.5cm of root, fill=blue!10] (t2) {Paris};

            \node[invalid, left=0.6cm of t1] (inv1) {London};

            % Level 2
            \node[trie, below=1.2cm of t2, fill=blue!10] (t3) {\_is};
            \node[trie, right=0.8cm of t2] (t4) {\_has};

            % Level 3
            \node[trie, below=1.2cm of t3, fill=blue!10] (t5) {\_in};
            \node[leaf, below=1.2cm of t4] (t6) {$\langle$/par$\rangle$};

            % Level 4
            \node[leaf, below=1.2cm of t5, fill=blue!10] (t7) {France};

            % Edges
            \draw[invalidarrow] (root) -- node[above left, sloped, font=\scriptsize, gray] {$P=0$} (inv1);
            \draw[arrow] (root) -- node[above left, font=\scriptsize] {"$\langle$par"} (t1);
            \draw[arrow] (root) -- node[above right, font=\scriptsize] {"Par"} (t2);

            \draw[arrow] (t2) -- node[left, font=\scriptsize] {" is"} (t3);
            \draw[arrow] (t2) -- node[right, font=\scriptsize] {" has"} (t4);

            \draw[arrow] (t3) -- node[left, font=\scriptsize] {" in"} (t5);
            \draw[arrow] (t4) -- node[right, font=\scriptsize] {" $\langle$/"} (t6);

            \draw[arrow] (t5) -- node[left, font=\scriptsize] {" Fran"} (t7);

            % Annotations
            \node[above=0.3cm of root, font=\footnotesize\bfseries] {Beam Search Hypothesis Path};
        \end{tikzpicture}
    }
    \caption{Visualization of KG-Trie constrained generation.}
    \label{fig:kg_trie_search}
\end{figure}

% -----------------------------------------------------------------------------
\section{Retrieval-Augmented Generation}
\label{sec:rag}
% -----------------------------------------------------------------------------

Retrieval-augmented generation (RAG), introduced by \citet{lewis2020rag}, is the dominant paradigm for grounding LLM outputs in external knowledge. A dense retriever selects relevant passages from an external source based on embedding similarity, and a reader LLM generates the answer conditioned on the retrieved context and the question. Fusion-in-Decoder~\citep{izacard2021leveraging} extends this by encoding multiple retrieved passages independently and fusing them in the decoder, improving coverage.

\subsection{KG-Augmented Retrieval}

Several works augment RAG with KG-structured context. \citet{he2021multihop} retrieve KG subgraphs and use attention over retrieved facts to guide multi-hop reasoning. \citet{yasunaga2021qagnn} combine retrieved text passages and KG neighborhoods via graph neural networks. \citet{shi2021kgpath} retrieve KG paths as natural language context strings, providing path-shaped input to the reader. G-Retriever~\citep{he2024gretriever} applies RAG specifically to textual graph understanding, retrieving graph-structured context for QA tasks.

\subsection{Advanced RAG Variants}

More recent RAG variants address the timing and precision of retrieval. Self-RAG~\citep{asai2024selfrag} inserts reflection tokens that allow the model to dynamically decide when retrieval is needed and whether retrieved passages support the claim being generated, improving retrieval precision and citation accuracy. FLARE~\citep{jiang2023flare} uses the model's next-token prediction confidence as a retrieval trigger, retrieving fresh context when confidence falls below a threshold. IRCoT~\citep{trivedi2022ircot} interleaves retrieval with chain-of-thought steps, retrieving relevant context at each intermediate reasoning step and improving multi-hop accuracy substantially. GraphRAG~\citep{edge2024graphrag} constructs community-structured knowledge graphs from document corpora and retrieves community summaries for complex relational queries, explicitly leveraging graph structure for improved coverage.

\subsection{The Structural Limitation of Soft Grounding}

Despite their empirical improvements, all RAG approaches share a structural limitation: retrieved context is a soft influence on generation. The retriever determines what the LLM sees; it does not determine what the LLM is permitted to generate. Because the LLM's generation mechanism remains unchanged (the distribution over the full vocabulary is computed autoregressively), there exists a nonzero probability of generating any token at any step, including tokens that correspond to no verified KG fact. Structural faithfulness with probability one, which guarantees that every generated token corresponds to a verified fact, cannot be achieved under any architecture that modifies only the input context without modifying the token selection mechanism~\citep{luo2024gcr, geng2023gcd}.

\begin{figure}[H]
    \centering
    \resizebox{\textwidth}{!}{
        \begin{tikzpicture}[>=latex, font=\sffamily\footnotesize]
            % Styles
            \tikzstyle{block} = [draw, rectangle, rounded corners, minimum height=1cm, minimum width=2.5cm, align=center, fill=blue!5]
            \tikzstyle{data} = [draw=none, fill=gray!10, rounded corners, inner sep=5pt, align=center]
            \tikzstyle{redanno} = [text=red!80!black, font=\sffamily\footnotesize\bfseries, align=center]
            \tikzstyle{greenanno} = [text=green!60!black, font=\sffamily\footnotesize\bfseries, align=center]

            % LEFT PANEL: RAG
            \node[font=\sffamily\normalsize\bfseries] at (-3, 3.5) {RAG Pipeline};
            \node[data] (q_left) at (-3, 2.3) {Question $q$};
            \node[block, fill=orange!10] (ret) at (-5.5, 1.2) {Retriever \\ (Soft Context)};
            \node[data] (ctx) at (-5.5, -0.2) {Retrieved \\ Subgraph/Text};
            \node[block] (llm_left) at (-3, -1.5) {LLM \\ (Autoregressive)};

            \draw[->, thick] (q_left) -| (ret);
            \draw[->, thick] (ret) -- (ctx);
            \draw[->, thick] (ctx) |- (llm_left);
            \draw[->, thick] (q_left) -- (llm_left);

            \node[data] (out_left) at (-3, -3) {Any token $v \in \mathcal{V}$ \\ possible};
            \draw[->, thick] (llm_left) -- (out_left);

            \node[redanno] at (-3, -4.5) {Soft Influence:\\ Faithfulness Not Guaranteed};

            % RIGHT PANEL: Constrained Decoding
            \node[font=\sffamily\normalsize\bfseries] at (4, 3.5) {Constrained Decoding Pipeline};
            \node[data] (q_right) at (4, 2.3) {Question $q$ \\ \& Prefix $y_{<t}$};
            \node[block, fill=teal!10] (oracle) at (1.5, 1.2) {Constraint Oracle \\ (KG-Trie)};
            \node[block] (llm_right) at (6.5, 1.2) {LLM \\ (Raw Logits)};

            \node[data] (mask) at (1.5, -0.2) {Binary Mask \\ $\alpha_t \in \{0, -\infty\}$};
            \node[data] (logits) at (6.5, -0.2) {Logits \\ $P(v \mid y_{<t}, x)$};
            \node[draw, circle, inner sep=3pt, fill=blue!10] (mul) at (4, -0.2) {\textbf{+}};

            \draw[->, thick] (q_right) -| (oracle);
            \draw[->, thick] (q_right) -| (llm_right);
            \draw[->, thick] (oracle) -- (mask);
            \draw[->, thick] (llm_right) -- (logits);
            \draw[->, thick] (mask) -- (mul);
            \draw[->, thick] (logits) -- (mul);

            \node[data] (out_right) at (4, -3) {Only valid KG tokens \\ $v \in \mathcal{W}_{\text{val}}(t)$ possible};
            \draw[->, thick] (mul) -- (4, -1.2) -- (out_right);
            \node[anchor=west, font=\scriptsize] at (4.1, -1.5) {Masked Logits};

            \node[greenanno] at (4, -4.5) {Hard Constraint:\\ Structural Faithfulness Guaranteed};

            % Divider
            \draw[dashed, gray, thick] (-0.5, -5) -- (-0.5, 4);

        \end{tikzpicture}
    }
    \caption{Architectural comparison of standard Retrieval-Augmented Generation (left) and KG-constrained decoding (right).}
    \label{fig:rag_vs_constrained}
\end{figure}

% -----------------------------------------------------------------------------
\section{Semantic Relevance Scoring with Sentence Transformers}
\label{sec:sbert}
% -----------------------------------------------------------------------------

\subsection{Sentence Transformer Embeddings}

Sentence-BERT~\citep{reimers2019sbert} introduced a class of sentence encoders trained to produce dense vector representations in which semantically similar sentences cluster together and semantically dissimilar sentences are pushed apart. The SBERT architecture applies mean pooling over token representations from a BERT-like transformer encoder, producing a fixed-length embedding. Training uses a siamese network on labelled sentence pairs with a classification objective over entailment, neutral, and contradiction labels. The resulting embedding space supports cosine similarity as a reliable proxy for semantic alignment between sentences, without requiring task-specific fine-tuning at inference time.

\subsection{Applications in Retrieval and Graph Reasoning}

Sentence transformer similarity has been applied extensively in dense retrieval settings. \citet{karpukhin2020dpr} demonstrate that a bi-encoder architecture trained on question-passage pairs produces embeddings that outperform sparse retrieval (BM25) on open-domain QA. \citet{xiong2020adore} refine bi-encoder training, improving coverage of relevant passages. In the KGQA context, \citet{shi2021kgpath} and \citet{saxena2022kgqa} apply semantic similarity to rank candidate KG paths for retrieval, selecting high-scoring paths as soft context for the reader LLM. These approaches use cosine similarity as a soft ranking signal over candidate paths, selecting the top-$k$ for context provision.

\subsection{Encoder Architecture Trade-offs}

The choice of sentence encoder involves a quality-latency tradeoff that is particularly stringent in step-wise constrained decoding, where the encoder is invoked at each generation step. Larger encoders such as all-mpnet-base-v2 (110M parameters) and BGE-M3~\citep{chen2024bge} (568M parameters) achieve higher performance on semantic textual similarity benchmarks but incur substantially higher inference latency. Lexical approaches such as BM25 are fast but cannot capture semantic similarity between paraphrase-equivalent paths that share no surface tokens. The all-MiniLM-L6-v2 model~\citep{reimers2019sbert}, at 22M parameters with approximately 80ms per-query CPU latency, provides the optimal balance of semantic quality and inference speed for iterative step-wise applications.

% -----------------------------------------------------------------------------
\section{Related Works}
\label{sec:related}
% -----------------------------------------------------------------------------
Building on the foundational concepts of constrained decoding, retrieval-augmented generation, and semantic scoring, this section steps back to survey the broader landscape of knowledge-augmented reasoning. To illustrate the evolution of the field, the surrounding literature is grouped into thematic clusters: LLM reasoning enhanced by knowledge graph access, structured retrieval-based question answering, and format-constrained generation. Examining these groups helps clarify both the significant achievements of prior methods and the limitations they share.

\subsection{Semantic Parsing and Early Structured KGQA}

Before the widespread adoption of large language models, classical approaches primarily tackled KGQA through semantic parsing or structured representation learning. There are two mainstream approaches corresponding to this era: parsing the question into executable KG queries like SPARQL, and grounding question and KG representations into a common space for reasoning.

In the first category, early semantic parsing methods relied on pre-defined question templates to synthesize queries, requiring strong domain knowledge~\citep{berant2013,reddy2014,bast2015}. More recent methods apply deep learning techniques to automatically generate these executable logical forms~\citep{yih2015,lan2020sparql}. While these symbolic approaches promise high interpretability and exact reasoning, they suffer from significant real-world bottlenecks: they require expensive ground-truth logical forms for supervision, which are costly to obtain, and their performance severely degrades when the KG has missing links. Furthermore, model-generated queries are frequently non-executable due to minor syntax or semantic mismatches with the underlying graph schema, yielding no answers and lacking graceful fallback mechanisms~\citep{yu2023decaf,luo2024gcr}.

To alleviate the need for rigid logical forms, the second category of methods grounds text and graph representations into a shared continuous space to reason over. Some match question representations directly to KG facts~\citep{miller2016,xu2019} or KG structural representations~\citep{zhang2018,sen-etal-2023-knowledge}. Prominent examples include GraftNet~\citep{sun2018graftnet}, PullNet~\citep{sun2019pullnet}, and NSM~\citep{he2021nsm}, which enhance reasoning through deeply embedded graph-based reasoners. Incomplete graphs have also drawn attention, where missing links are inferred using KG embeddings~\citep{saxena2020}. While these structured techniques successfully side-step hallucination by operating natively over verified data, their expressive capability is limited compared to the generative flexibility and sophisticated multi-step reasoning offered by modern LLMs.

\subsection{Knowledge Graph-Enhanced LLM Reasoning}

More recent advances focus on the joint use of KGs and LLMs for KGQA, leveraging the sophisticated capabilities of LLMs for enhanced reasoning while mitigating hallucinations and knowledge gaps. Foundational studies initially explored eliciting reasoning via prompting alone, such as Chain-of-Thought (CoT)~\citep{wei2022cot}, Plan-and-solve~\citep{wang2023c}, DecomP~\citep{khot2023decomposed}, Self-consistency~\citep{wang2023-self-consistency}, and Tree-of-Thought (ToT)~\citep{yao2023-tree-of-thought}. Various studies also focus on fine-tuning LLMs on reasoning tasks to improve their abilities~\citep{hoffman2024}, with efforts extending to reinforcement learning for complex reasoning processes (e.g., OpenAI o1). However, the persistent risk of hallucinations and missing domain knowledge compromises the faithfulness of pure generative reasoning, prompting the field to augment LLMs directly with KGs.

A large body of work explores how LLMs can be enhanced with KG access to guide derivations. Several methods treat LLMs as agents interacting with environments or KGs, such as ReACT~\citep{yao2022react}, StructGPT~\citep{jiang2023structgpt}, and ToG~\citep{sun2024tog}, which prompts the LLM to explore relevant facts hop-by-hop using beam search on the KG. Others retrieve subgraphs or paths for intermediate reasoning supervision: KD-CoT~\citep{wang2023knowledgedrivencotexploringfaithful} and RR~\citep{he2022rethinking} retrieve facts to guide generative plans, while \citet{jin2024graphcot} propose Graph Chain-of-Thought (GraphCoT) to explicitly retrieve graph neighbourhood information at every intermediate CoT step.

Specialised retrievers and graph encoders have also been developed to bridge text and graph modalities. Approaches like GNN-RAG~\citep{mavromatis2024gnnrag} and GFM-RAG~\citep{luo2025gfmrag} adopt graph neural networks to effectively capture structures and retrieve from KGs. \citet{luo2024rog} introduce Reasoning on Graphs (RoG), a planning-retrieval-reasoning framework that retrieves reasoning paths from KGs to conduct faithful reasoning. StructLM~\citep{zhuang2024structlm} fine-tunes generalist models over massive structured datasets to fundamentally augment grounding capabilities. DECAF~\citep{yu2023decaf} combines semantic parsing and LLM reasoning to jointly generate answers. While these approaches push the boundary of performance, they remain heavily reliant on instruction-following abilities (challenging for compact models) or computationally expensive fine-tuning.

Instead of strictly retrieving limited facts, methods such as Decoding on Graphs (DOG)~\citep{li2024dog} employ a general graph-aware decoding approach that uses the entire question graph natively without external retrievers, thus enabling reasoning over complete knowledge. Tree-of-Traversals~\citep{markowitz-etal-2024-tree} similarly offers a zero-shot algorithm motivated by ToT to explore alternative solution structures.

Despite these advancements, many of these approaches apply KG knowledge as an advisory rather than hard structural constraint. For instance, \citet{jiang2022unikgqa} propose UniKGQA, unifying retrieval and reasoning for multi-hop KGQA through a single model with a shared parameter space. By tying the retrieval and reasoning objectives to a common representation, UniKGQA achieves strong performance. The shared representation, however, means that errors in retrieval directly bias reasoning and vice versa, with no architectural separation between the two failure modes.

\citet{sun2018openke} develop OpenKE, an open-source platform for knowledge embedding that provides entity and relation representations learned from KG structure. These embeddings underpin downstream KGQA systems by providing semantically meaningful initializations for entity and relation tokens. A key limitation is that embedding-based representations are trained on fixed KG snapshots and do not generalise to new entities or relations that appear after training. Moreover, embedding similarity does not constitute a hard faithfulness guarantee.

\citet{zhang2022subgraph} propose a subgraph-based KGQA approach that retrieves a question-relevant subgraph and performs message passing over it to propagate relational evidence before the LLM generates an answer. This approach substantially reduces the search space by focusing on a localised subgraph, improving efficiency. Its limitation is that the subgraph boundary is determined heuristically based on entity proximity, and correct answer entities that are far from the question entity in graph distance may be excluded from the subgraph entirely, creating a recall ceiling that is invisible to the answering module.

\citet{wang2021kepler} present KEPLER, a model that jointly optimises knowledge embedding and masked language modelling objectives within a single transformer, integrating factual KG knowledge directly into LLM parameters during pretraining. KEPLER improves entity representations and reduces hallucination on entity-centric tasks compared to models without KG pretraining. The limitation is that KG knowledge is internalised as parametric memory rather than enforced at generation time: KEPLER-pretrained models remain subject to the same autoregressive generation mechanism and can still hallucinate facts, particularly for long-tail entities underrepresented in the KG-augmented pretraining corpus.

\citet{sun2019pullnet} introduce PullNet, an iterative graph construction and reasoning framework for multi-hop QA over heterogeneous text-and-KG sources. PullNet iteratively pulls relevant entities and facts from both corpora, building a question-specific graph over which inference is performed. This iterative construction improves coverage over fixed-subgraph retrieval approaches. However, because the graph is constructed from both text and KG, it conflates verified KG facts with unverified textual claims, and the downstream reasoning module cannot distinguish between the two. The faithfulness guarantee provided by the KG is therefore diluted by the presence of textual evidence with no formal verification status.

\citet{he2022rethinkingKG} conduct an empirical rethinking of KG-augmented LLM reasoning, systematically comparing a range of KG integration strategies across multiple benchmarks. Their analysis reveals that na\"{i}ve KG injection (concatenating retrieved facts into the LLM prompt) often underperforms fine-tuned models, and that the quality of KG evidence retrieval is the primary performance bottleneck across all strategies. This finding motivates more sophisticated retrieval mechanisms but does not address the structural faithfulness problem: even with oracle retrieval, the LLM remains free to ignore or contradict the provided facts during generation.

\citet{mavromatis2022rearev} propose ReaRev, a reasoning revision framework for KGQA that applies iterative reasoning and revision over retrieved KG evidence, using a graph neural network to propagate and revise entity scores across reasoning steps. ReaRev achieves competitive performance on CWQ by allowing the model to revise earlier incorrect entity assignments as more evidence accumulates. The fundamental limitation is that the revision procedure operates over soft entity scores rather than hard path commitments: the model does not generate a verifiable reasoning chain, and the final entity selection is made by argmax over a distribution rather than by following a path that can be independently verified against the KG.

These works collectively demonstrate the value of explicit KG access for improving LLM reasoning quality and interpretability. Their shared limitation is that KG access is advisory rather than constraining: the LLM attends to retrieved KG context but is free to generate beyond it. Across all approaches reviewed, the reasoning chains produced may be plausible and KG-informed without being guaranteed to correspond to verified KG paths. The absence of a hard structural guarantee is not merely a theoretical concern: \citet{luo2024gcr} measure this gap directly, finding that even the strongest KG-enhanced approaches produce hallucinated reasoning steps in approximately one-third of cases, establishing a systematic upper bound on what soft KG grounding can achieve.

\subsection{Retrieval-Based KGQA}

Earlier KGQA approaches relied on structured retrieval over the knowledge graph rather than LLM generation. These methods avoid LLM hallucination by never invoking a free-form generative model, operating instead within the structured boundaries of the KG. The following works span the primary mechanisms in this family, from semantic parsing and embedding-based retrieval to graph neural network propagation and path-level supervision.

\citet{yih2016websqp} establish the WebQSP benchmark and introduce a semantic parsing approach that maps natural language questions to SPARQL queries executed against Freebase. By constraining the answer derivation to formally verified SPARQL execution, this approach provides exact symbolic faithfulness. Its core limitation is the brittleness of the parsing step: small surface variations in question phrasing can produce incorrect parses that yield structurally valid but semantically incorrect SPARQL queries. The approach also requires exhaustive annotation of SPARQL queries for training, which is costly and difficult to scale to complex compositional questions.

\citet{talmor2018cwq} introduce ComplexWebQuestions and a corresponding compositional semantic parsing approach that decomposes complex questions into simpler sub-questions before generating SPARQL. This decomposition strategy extends the coverage of semantic parsing to compositional queries but does not resolve the brittleness problem: decomposition errors compound across sub-questions, and the final SPARQL query may be structurally valid while answering a subtly different question than the one posed.

\citet{he2021multihop} propose a multi-hop KGQA method that learns intermediate supervision signals to guide path selection through the KG, showing that explicit intermediate supervision at each hop substantially improves performance at greater reasoning depths. By training the model to predict correct intermediate entities, this approach reduces the combinatorial difficulty of multi-hop search. Its limitation is that intermediate supervision signals are derived from annotated answer paths, which are expensive to obtain and may not be available for every benchmark or KG. Moreover, the approach is constrained to a fixed maximum hop depth and does not adapt its reasoning depth dynamically based on question complexity.

\citet{yasunaga2021qagnn} propose QA-GNN, combining retrieved text passages and KG neighbourhood entities in a joint graph neural network that propagates information between passage token nodes and KG entity nodes. By jointly reasoning over both modalities, QA-GNN improves performance on questions that require combining textual and structured evidence. The joint graph conflates verified KG facts with unverified textual claims in the same propagation mechanism, meaning that noisy or incorrect text passages can corrupt entity scores derived from KG structure. The approach also requires constructing a joint graph at inference time for each question, incurring significant preprocessing overhead.

\citet{shi2021kgpath} retrieve KG reasoning paths by computing semantic similarity between the question and natural language descriptions of candidate paths, providing the top-$k$ highest-scoring paths as context strings to a reader LLM. This retrieval-plus-reading architecture separates the structured retrieval step from the natural language generation step. The limitation is that the ranking function uses soft cosine similarity over surface-form path descriptions, which may assign high scores to paths that are lexically similar to the question but semantically incorrect. The reader LLM, receiving paths as plain text, has no mechanism to verify that the provided paths are actual KG edges.

\citet{saxena2022kgqa} apply entity and relation embeddings to score and select multi-hop KG paths for KGQA without invoking an LLM generator. Entity and relation representations are learned from KG structure using embedding objectives, and multi-hop path scores are computed by composing relation embeddings. This approach is fully contained within the structured KG and avoids generative hallucination entirely. However, embedding-based path scoring cannot capture the full semantic complexity of natural language questions: compositional questions with negation, comparison, or conjunction are difficult to encode faithfully as compositions of relation embeddings, creating a coverage ceiling on complex question types.

\citet{lan2022kgqa} provide a comprehensive survey of KGQA approaches, systematically analysing the trade-offs between semantic parsing, embedding-based, and retrieval-based methods across benchmarks. Their analysis reveals that retrieval-based approaches consistently struggle with tail entities and uncommon relation types that are underrepresented in training data, and that no single retrieval mechanism dominates across all question types and hop depths. This survey motivates the turn toward LLM-based approaches as a more flexible alternative.

\citet{sun2019pullnet} develop PullNet, which iteratively pulls relevant KG facts and text passages into a dynamically constructed evidence graph, performing inference over this graph to answer multi-hop questions. The iterative construction allows PullNet to retrieve evidence at multiple depths without fixing the hop depth in advance. Its limitation is that the evidence graph mixes KG and text sources without distinguishing their reliability, and the iterative pull procedure introduces substantial latency that scales with question complexity.

\citet{das2017go} introduce ``Go for a Walk and Arrive at the Answer,'' an end-to-end differentiable path-finding approach that trains a policy to traverse the KG from a question entity to the answer entity. The policy is trained with reinforcement learning using answer correctness as reward. This approach directly models the reasoning process as graph traversal and produces interpretable paths. Its limitation is that reinforcement learning from sparse answer-correctness rewards is highly unstable during training, requiring careful reward shaping. The policy also generalises poorly to question types not well-represented in the training distribution.

\citet{zhang2022subgraph} propose subgraph-based reasoning that extracts a question-relevant subgraph from the full KG and performs message passing over it before selecting the answer entity. By restricting reasoning to a local subgraph, this approach dramatically reduces the search space and inference cost. The subgraph boundary is determined by entity proximity to the question entity, which creates a recall ceiling: answer entities that are topologically distant from the question entity in the KG may be excluded from the subgraph even when they are the correct answer. This limitation is structurally linked to the hop-depth assumptions built into subgraph extraction heuristics.

These retrieval-based approaches avoid LLM hallucination by operating within the structured KG rather than generating freely. Their shared limitation is expressiveness: they are typically constrained to fixed path structures, fixed hop depths, or fixed retrieval budgets, and do not leverage the flexible multi-step reasoning capability of modern LLMs. The transition to LLM-based approaches in more recent work improved naturalness and coverage on complex compositional questions, but at the cost of the structural faithfulness guarantees that retrieval-based methods implicitly provide.

\subsection{Format-Constrained and Entity-Constrained Generation}

A distinct line of research addresses factual faithfulness by imposing hard constraints on what a generative model is permitted to produce, independently of the input context. Unlike retrieval-based approaches that restrict the information available to the model, these works modify the generation mechanism itself. The following cluster traces the development of this paradigm from early lexical constraints through grammar-based and automaton-based frameworks.

\citet{decao2021autoregressive} introduce autoregressive entity retrieval (GENRE), constraining generation to a finite set of entity names using a trie constructed over all known entity surface forms. By restricting beam search to valid trie paths, GENRE guarantees that every generated token sequence corresponds to a real entity in a fixed vocabulary. This approach eliminates hallucination of entity names and achieves strong performance on entity disambiguation and linking tasks. Its limitation for KGQA is that constraining to entity names does not enforce path-level faithfulness: generating a real entity name at each step does not ensure that consecutive entities are connected by a verified KG edge. A sequence of individually valid entity names may constitute a reasoning chain that has no corresponding path in the KG.

\citet{geng2023gcd} generalise entity-level trie constraints to arbitrary context-free grammar constraints in Grammar-Constrained Decoding (GCD). GCD uses an Earley parser as the constraint oracle, admitting only tokens that extend the current partial string as a valid grammar prefix. GCD additionally introduces input-dependent grammars parameterised by the input question, allowing the valid token set to depend on question content. GCD establishes that grammar constraints are computationally tractable and empirically improve structured output quality across multiple tasks. The limitation is that a context-free grammar can enforce syntactic or format correctness but cannot encode factual membership in a knowledge graph: a string may conform perfectly to a grammar specification while containing entirely hallucinated entities or relations.

\citet{willard2023outlines} address the computational cost of grammar-constrained decoding with an efficient finite-state machine (FSM) formulation. By precomputing an index from FSM states to valid vocabulary subsets, the valid token set can be retrieved in amortised constant time at each generation step, making constrained decoding practical at inference speed for production systems. This formulation reduces the per-step constraint overhead from the Earley parsing cost (which is cubic in input length) to effectively $O(1)$ amortised lookup. The limitation remains the same as GCD: the FSM encodes structural or format correctness, not factual correctness. An FSM-constrained system can still generate structurally valid outputs that are factually incorrect with respect to any external knowledge source.

\citet{deutsch2019constrained} propose a general algorithm for constrained sequential inference applicable across any structured prediction task, framing constraint enforcement as inference in a product of finite-state automata. This framework unifies a broad range of constraint types under a common algorithmic schema and demonstrates that hard constraint enforcement is compatible with a variety of sequence models. Its limitation for KGQA is that the constraints it enforces are defined over output string structure rather than over external knowledge: the automata encode properties of the output form, not membership in a KG edge set.

\citet{poesia2022synchromesh} introduce Synchromesh, a constrained generation system for program synthesis that uses a completion engine to enforce syntactic validity of generated code. By grounding constraint enforcement in a language-specific parser, Synchromesh ensures that all generated programs are syntactically valid. This work establishes that constraint oracles can be derived from formal language specifications beyond context-free grammars. Applied to KGQA, an analogous approach would enforce syntactic validity of SPARQL queries; however, a syntactically valid SPARQL query may still reference entities or relations absent from the target KG, producing an empty result set rather than a faithfully grounded answer.

\citet{scholak2021picard} present PICARD, a method for parsing SQL queries with constrained autoregressive inference. PICARD uses an incremental parser to reject tokens that would make the partial output unparseable or semantically invalid with respect to a database schema, thereby constraining generation to schema-valid SQL. PICARD demonstrates that schema-grounded constraints substantially reduce invalid query generation in text-to-SQL tasks. The limitation for KG settings is that SQL schema constraints enforce structural consistency with a fixed schema, not factual path traversal through a dynamic relational graph. A schema-valid query may still produce an incorrect answer if the underlying reasoning is unsound.

\citet{anderson2017guided} introduce Guided Open Vocabulary Image Captioning (GOVIC), an early demonstration of constrained beam search for lexically-constrained generation in image captioning, enforcing inclusion of specified words in the output. While the domain differs from KGQA, this work establishes constrained beam search as a practical mechanism for enforcing lexical constraints during generation and demonstrates the feasibility of hard inclusion constraints. Its limitation is that lexical constraints enforce presence of specific surface forms without enforcing their factual correctness or logical coherence with the rest of the generated output.

\citet{hokamp2017lexically} introduce constrained decoding via a grid beam search algorithm that enforces specified lexical constraints in neural machine translation. Grid beam search maintains separate hypothesis sets for each subset of satisfied constraints and merges them to produce outputs satisfying all constraints. This algorithm demonstrates that complex constraint sets can be enforced in polynomial time during beam search. The limitation is that the constraint set must be specified in advance and is fixed throughout generation; there is no mechanism for the constraint oracle to consult an external knowledge source at each step to determine which tokens are valid continuations.

\citet{lu2021neurologic} propose NeuroLogic Decoding, a decoding algorithm that enforces propositional logic constraints over lexical items during generation, extending the class of enforceable constraints beyond simple inclusion to disjunctions, conjunctions, and negations. NeuroLogic demonstrates that logically complex constraints can be enforced efficiently during beam search without prohibitive computational overhead. For KGQA applications, the limitation is that propositional logic constraints over lexical items still operate at the surface-form level: a NeuroLogic constraint can require that a specific entity name appears in the output, but cannot verify that this entity is connected to the preceding entity by a verified KG edge.

\citet{zhang2023tractable} analyse the computational tractability of constrained decoding under various constraint classes, establishing theoretical bounds on the complexity of constraint enforcement as a function of grammar type and vocabulary size. Their analysis confirms that finite-state constraints admit efficient enforcement via precomputed indices, while context-sensitive constraints require per-step evaluation that may be prohibitively slow. This theoretical result directly motivates the design choice in KG-constrained systems of representing valid paths as tries (finite-state structures) rather than as arbitrary graph reachability queries.

These works collectively establish that hard constraints on generation are computationally tractable and empirically beneficial across multiple domains. Their shared limitation for KGQA is that they enforce output \textit{format} rather than \textit{factual content}: constraining generation to grammatically valid strings, schema-valid queries, or lexically specified surface forms does not ensure that the generated strings correspond to verified KG paths. The extension from format constraints to KG path constraints requires a constraint oracle that encodes the relational structure of the knowledge graph itself --- mapping valid token sequences not to grammar derivations but to traversable edges in a specific, dynamic knowledge base.

\subsection{Semantic Similarity in Graph Reasoning}

Several works apply semantic similarity signals to guide reasoning over knowledge graphs. \citet{shi2021kgpath} rank candidate KG paths by semantic similarity between the question and path descriptions, retrieving high-scoring paths as evidence. \citet{saxena2022kgqa} learn embeddings for KG entities and relations that support similarity-based multi-hop path selection. \citet{he2021multihop} use attention-guided traversal over retrieved KG subgraphs, with attention weights reflecting the relevance of each subgraph region to the question. PathRetriever~\citep{asai2019learning} learns to retrieve reasoning paths from evidence graphs by scoring paths using a recurrent model over their node sequences.

These works demonstrate that semantic similarity can effectively differentiate relevant from irrelevant KG paths in a retrieval setting. The key architectural distinction is that semantic similarity is applied as a \textit{soft ranking signal}: candidate paths are scored and the highest-scoring ones are provided to the LLM as context or used to select among generated candidates. No existing work applies semantic similarity as a \textit{hard binary admission criterion} that determines which tokens the LLM is permitted to generate at each decoding step. The distinction between soft retrieval signals and hard generation constraints is significant: the former influences what the LLM sees, while the latter determines what the LLM can produce.

\subsection{Hallucination Mitigation in Knowledge-Intensive Tasks}

Beyond the constrained decoding paradigm, a range of approaches attempt to reduce hallucination in knowledge-intensive tasks. Retrieval-augmented approaches (including RAG~\citep{lewis2020rag}, Self-RAG~\citep{asai2024selfrag}, FLARE~\citep{jiang2023flare}, IRCoT~\citep{trivedi2022ircot}, and GraphRAG~\citep{edge2024graphrag}) address the problem by providing relevant external context, improving factual coverage without modifying the generation mechanism. \citet{wang2023selfconsistency} improve consistency by majority-voting over multiple CoT chains. \citet{huang2024selfcorrect} study whether LLMs can self-correct reasoning errors, finding that self-correction is generally ineffective without external feedback, reinforcing that internal parametric knowledge cannot reliably detect its own errors.

Post-hoc verification approaches~\citep{nakano2021webgpt, gao2023rarr} check generated claims against external sources after generation and revise or flag unsupported statements. While effective at identifying errors, these approaches require a separate verification pass and do not prevent hallucination during generation. \citet{agrawal2024mindful} systematically analyse failure modes in RAG systems, finding that even with relevant context provided, LLMs regularly produce statements unsupported by the retrieved evidence.

\subsection{Concurrent Work}

RouterKGQA~\citep{yuan2026routerkgqa} is the work most closely related to this research in terms of shared components. RouterKGQA routes KGQA queries between a specialised KG reasoning model and a general LLM based on question complexity, and applies SBERT-based semantic similarity (using the nomic-embed-text-v1 encoder) to filter candidate reasoning paths. It achieves competitive performance with reduced inference cost, averaging approximately 1.15 LLM calls per question.

The key architectural distinction is the level at which semantic scoring is applied. RouterKGQA scores complete reasoning paths after they have been generated, using similarity to filter among candidates retrospectively. This approach does not modify the token selection mechanism during generation: the LLM generates freely, and semantic similarity is applied post-hoc to select among the outputs. Applying semantic similarity as a hard admission criterion at the token constraint level — determining which tokens the LLM is permitted to generate at each decoding step rather than which outputs to keep after generation — represents a qualitatively different architectural intervention that RouterKGQA does not explore. RouterKGQA also does not provide a metric for constraint permissiveness and does not measure the degree to which its candidate sets contain structurally valid but semantically irrelevant paths.

% -----------------------------------------------------------------------------
\section{Synthesis: Three Unresolved Gaps}
\label{sec:synthesis}
% -----------------------------------------------------------------------------

The preceding review reveals that two distinct research trajectories (constrained decoding for structural faithfulness, and semantic similarity for relevance-guided reasoning) have developed largely in parallel without being integrated at the constraint-oracle level. Constrained decoding approaches~\citep{luo2024gcr, li2024dog} achieve structural faithfulness but construct their constraint oracles without reference to question semantics or generation context, producing valid token sets that contain large numbers of irrelevant paths. Semantic similarity approaches~\citep{shi2021kgpath, saxena2022kgqa, yuan2026routerkgqa} improve relevance but operate as soft signals that influence retrieval or post-hoc selection without modifying the generation mechanism.

Three specific gaps emerge from this review.

\textbf{Gap 1 -- The Permissiveness Problem.} Existing KG-constrained decoding frameworks construct their constraint oracles as functions of graph structure and question entities alone, without reference to question semantics or the evolving generation state. For multi-hop questions, this produces valid token sets containing thousands of structurally correct but semantically irrelevant paths at every generation step. The LLM must discriminate among these paths using its own parametric knowledge, reintroducing the hallucination risk that constrained decoding was designed to eliminate~\citep{luo2024gcr, li2024dog}.

\textbf{Gap 2 -- The Static-Dynamic Efficiency Tradeoff.} GCR's precomputed static trie is computationally efficient but semantically blind and identical at every generation step. DoG's step-wise topological expansion introduces dynamism at the cost of per-step trie reconstruction and without semantic filtering~\citep{li2024dog, willard2023outlines}. No existing framework achieves semantic responsiveness (where the valid set narrows progressively as reasoning commits to a specific direction) while preserving the efficiency of precomputed constraint structures.

\textbf{Gap 3 -- The Absence of a Constraint Quality Metric.} No existing work provides a metric that quantifies constraint permissiveness independently of downstream answer accuracy. Without such a metric, it is impossible to measure progress on reducing irrelevant paths in the valid set, compare frameworks on constraint quality independently of model quality, or diagnose permissiveness as a failure mode in deployed systems~\citep{luo2024gcr, li2024dog}.

\begin{table}[H]
    \centering
    \caption{Summary of key related works and gap coverage.}
    \label{tab:related_summary}
    \small
    \resizebox{\textwidth}{!}{%
        \begin{gridtabular}{l|l|l|l}
            \gr{\textbf{Work}               & \textbf{Approach}                & \textbf{Structural faithfulness} & \textbf{Semantic conditioning} }
            \gr{\citet{luo2024gcr}          & Static KG-Trie decoding          & Yes (100\%)                      & No                              }
            \gr{\citet{li2024dog}           & Step-wise topological expansion  & Yes (100\%)                      & No                             }
            \gr{\citet{yuan2026routerkgqa}  & Routing + post-hoc filtering     & Probabilistic                    & Yes (answer-level)              }
            \gr{\citet{luo2024rog}          & Reasoning path planning          & No                               & Partial                         }
            \gr{\citet{jiang2023structgpt}  & LLM over structured data         & No                               & Partial                         }
            \gr{\citet{jin2024graphcot}     & Graph-augmented CoT              & No                               & Partial                         }
            \gr{\citet{jiang2022unikgqa}    & Unified retrieval and reasoning  & No                               & Partial                         }
            \gr{\citet{wang2021kepler}      & KG-pretrained language model     & No                               & Partial                         }
            \gr{\citet{mavromatis2022rearev}& Iterative reasoning revision     & No                               & Partial                         }
            \gr{\citet{he2021multihop}      & KG subgraph attention            & No                               & Partial                         }
            \gr{\citet{yasunaga2021qagnn}   & Text-KG graph neural network     & No                               & Partial                         }
            \gr{\citet{das2017go}           & RL-based KG path traversal       & No                               & Soft                            }
            \gr{\citet{saxena2022kgqa}      & Embedding-based path selection   & No                               & Soft                            }
            \gr{\citet{shi2021kgpath}       & Path retrieval by similarity     & No                               & Soft (retrieval)               }
            \gr{\citet{asai2024selfrag}     & Retrieval with reflection        & No                               & Soft                            }
            \gr{\citet{edge2024graphrag}    & Community-structured retrieval   & No                               & Soft                           }
            \gr{\citet{geng2023gcd}         & Grammar-constrained decoding     & Format only                      & No                              }
            \gr{\citet{willard2023outlines} & FSM-indexed constrained decoding & Format only                      & No                              }
            \gr{\citet{decao2021autoregressive} & Entity trie-constrained decoding & Entity names only             & No                              }
            \gr{\citet{scholak2021picard}   & Schema-constrained SQL parsing   & Schema only                      & No                              }
            \gr{\citet{lu2021neurologic}    & Logic-constrained decoding       & Format only                      & No                              }
        \end{gridtabular}
    }
\end{table}

These three gaps collectively identify an open problem at the intersection of the two research trajectories reviewed in this chapter: integrating semantic relevance scoring directly into the constraint oracle, conditioning the valid token set jointly on the knowledge graph, the question semantics, and the evolving generation state.
