% FrontMatter/abstract.tex - Project abstract

\begin{abstract}
	\addchaptertocentry{\abstractname}
	\vspace{0.5cm}

	\addcontentsline{toc}{chapter}{ABSTRACT}
	Large language models (LLMs) are capable of multi-hop reasoning but remain prone to hallucination. Knowledge graph-constrained decoding mitigates this by strictly restricting output tokens to valid paths within a knowledge graph (KG), ensuring absolute structural faithfulness unlike standard soft-grounding approaches (e.g., Retrieval-Augmented Generation). However, leading constrained paradigms—like Graph-Constrained Reasoning (GCR)—construct constraints solely from graph structure, ignoring the question's semantics and the evolving generation state.

	This causes the \emph{permissiveness problem}: at every decoding step, the allowed valid token set contains thousands of structurally accessible but contextually irrelevant paths. This flood of valid tokens forces the LLM to rely on the very parametric biases that constrained decoding was designed to suppress.

	To resolve this, we propose Dynamic Context-Aware Tries (DCA-Trie), a novel framework that conditions the valid tokens jointly on the knowledge graph, the input question, and the partially generated reasoning chain. It upgrades the constraint oracle to an active semantic state $f(\mathcal{G}, E_q, q, y_{<t})$ by using a pretrained relevance scorer to filter candidate KG paths. Two architectures are developed: static semantic filtering applied at trie construction (DCA-Trie v1), and dynamic, step-wise bounding conditioned on the evolving generation (DCA-Trie v2).

	Both variants are evaluated on the WebQSP and ComplexWebQuestions datasets over Freebase. Alongside standard metrics (Hits@1, F1), we introduce the Semantic Irrelevance Ratio (SIR)—a novel metric to uniquely isolate and quantify constraint permissiveness. Results demonstrate that DCA-Trie successfully eliminates semantically blind branches while preserving 100\% structural faithfulness, charting a direct path toward more robust and efficient constrained reasoning.
\end{abstract}
